`est_seroincidence()` maximizes the log-likelihood numerically with
`stats::nlm()`. Exit codes 3 to 5 flag a possible non-convergence, so they are
worth checking before trusting any of the estimates:

```{r}
#| label: check-convergence-codes
problem_ids <- ests_summary |>
  tibble::as_tibble() |>
  dplyr::filter(nlm.convergence.code > 2) |>
  dplyr::select(sample_size, lambda.sim, cluster)

nrow(problem_ids)
```

::: notes
Remedies, all passed through to `nlm()`:

* code 3: decrease `stepmin`
* code 4: increase `iterlim`
* code 5: increase `stepmax`

Pass them with `num_cores = 1`. `est_seroincidence_by()` currently drops
these arguments in its parallel branch
([#629](https://github.com/UCD-SERG/serocalculator/issues/629)), so under
parallelism they are silently ignored.
:::

```{r}
#| label: fig-problem-strata
#| fig-cap: >
#|   Log-likelihood curves for any clusters whose optimizer reported a
#|   possible convergence problem.
#| fig-height: 7
if (nrow(problem_ids) > 0) {
  refit_strata(
    pop_data = sim_df,
    strata_ids = problem_ids,
    sr_params = sim_curve_params,
    noise_params = sim_noise_params,
    curve_strata_varnames = NULL,
    noise_strata_varnames = NULL,
    antigen_isos = antibodies,
    num_cores = 1
  ) |>
    autoplot(log_x = TRUE)
}
```
