In many survey designs, observations are clustered (e.g., multiple individuals from the same household, school, or geographic area).
Observations within the same cluster are often more similar to each other than to observations from different clusters, which violates the independence assumption of @prp-full-sample-likelihood.
The estimate $\hat\lambda$ itself remains usable, but the inverse-information standard error of @thm-mle-asymptotic-variance no longer applies to it.

### Why clustering matters

When observations are clustered:

* Individuals within the same cluster share common exposures or characteristics
* Standard errors that ignore clustering are usually **too small** (anti-conservative)
* Confidence intervals are then **too narrow**
* p-values are then **too optimistic**

Usually rather than always: the direction follows the sign of the within-cluster correlation, which is positive in most survey designs but need not be.

### Cluster-robust variance estimation

To account for within-cluster correlation, `serocalculator` implements the **sandwich estimator** --- also known as the Huber-White robust variance estimator [@Huber1967; @White1980], generalized to clustered/grouped data by @LiangZeger1986.

::: {#def-cluster-robust-variance}

#### Sandwich variance estimator

Let $C$ be the total number of non-overlapping clusters in the sample,
and let $H = \hess(\hat\lambda)$ be the Hessian (@def-hessian) at the
MLE --- `serocalculator` obtains it from `stats::nlm()`'s own
`hessian` output.
Write $U_i$ for observation $i$'s score contribution (@def-score) --- the
gradient of its own log-likelihood $\llik_i$ (@prp-full-sample-likelihood)
--- and $U_c$ for the total over the observations in cluster $c$:

$$
\begin{aligned}
U_i &= \nabla_\lambda \llik_i(\lambda)
\\
U_c &= \sum_{i \in c} U_i
\end{aligned}
$$

where $\nabla_\lambda$ denotes the gradient with respect to $\lambda$.
These are gradients of the same $\llik_i$ that
@prp-full-sample-likelihood defines: the full-sample log-likelihood
sums $\llik_i$ over every observation, and the sandwich accumulates
their gradients cluster by cluster.
The **cluster-robust variance-covariance matrix** for the parameter
estimates is then

$$
\begin{aligned}
V_{\text{robust}} &= H^{-1} B H^{-1}
\\
B &= \sum_{c=1}^C U_c U_c^T
\end{aligned}
$$

with $B$ the "meat" of the sandwich.
:::

::: notes
The sign convention does not matter here: $H$ appears twice, so
substituting the observed information $\oinf(\hat\lambda) = -H$
(@def-observed-information) leaves $V_{\text{robust}}$ unchanged.

The sandwich generalizes @thm-mle-asymptotic-variance rather than
replacing it.
When the model is correctly specified and observations really are
independent, $B$ estimates the same information as $-H$ does, and
$H^{-1} B H^{-1}$ collapses to $\sb{\oinf(\hat\lambda)}^{-1}$ --- the
ordinary standard error.
Clustering is what drives the two apart, and the gap between them is
what the correction recovers.
:::

### Multi-way clustering

When `cluster_var` names more than one variable,
`serocalculator` does not collapse them into a single interaction.
Instead it applies the inclusion-exclusion identity of
Cameron, Gelbach, and Miller (2011) [@Cameron_2011]
for multi-way cluster-robust variance estimation.

::: {#def-multiway-clustering-variance}

#### Multi-way clustering-robust variance

Let $\mathcal{K} = \{k_1, \dots, k_p\}$ denote the cluster variables.
For a non-empty subset $S \subseteq \mathcal{K}$,
let $V_S$ be the one-way sandwich variance $V_{\text{robust}}$
(@def-cluster-robust-variance),
computed by clustering on the interaction of the variables in $S$.
The multi-way variance is the signed sum over every such subset:

$$V_{\text{multiway}} = \sum_{\emptyset \neq S \subseteq \mathcal{K}} (-1)^{|S| + 1} V_S$$

Computing every subset costs $2^p - 1$ one-way variance evaluations,
so this is intended for the small $p$ (2-3)
typical of nested survey designs.
:::

::: {#exm-two-way-clustering}

#### Two cluster variables

For two cluster variables ($p = 2$),
@def-multiway-clustering-variance reduces to the familiar
three-term formula:

$$V_{\text{multiway}} = V_1 + V_2 - V_{1 \cap 2}$$

where $V_1$ and $V_2$ cluster on each variable separately,
and $V_{1 \cap 2}$ clusters on their interaction,
the most granular grouping.
:::

### Small-sample correction

Two adjustments refine the one-way variance terms $V_S$ that feed
@def-multiway-clustering-variance.

::: {#def-finite-difference-score}

#### Finite-difference score approximation

`serocalculator` approximates each cluster's score contribution $U_c$
(@def-cluster-robust-variance) by a forward finite difference of the
cluster's log-likelihood,
$\ell_c(\lambda) = \sum_{i \in c} \log p(Y_i \mid \lambda)$:

$$U_c \approx \frac{\ell_c(\hl + \epsilon) - \ell_c(\hl)}{\epsilon}, \quad \epsilon = 10^{-6}$$
:::

::: {#def-cr1-correction}

#### CR1 small-sample correction

When `small_sample = "CR1"` (the default),
each one-way subset variance $V_S$
(@def-multiway-clustering-variance)
is inflated by the usual cluster degrees-of-freedom correction:

$$V_S^{\text{CR1}} = \frac{G_S}{G_S - 1} \, V_S$$

where $G_S$ is the number of unique clusters in subset $S$'s grouping.
:::

### Safeguards against invalid variance

::: {#def-variance-safeguards}

#### Flooring and fallback for the final variance

The inclusion-exclusion sum $V_{\text{multiway}}$
(@def-multiway-clustering-variance)
is not guaranteed to be non-negative.
This is a documented property of the Cameron-Gelbach-Miller estimator,
not a bug.
`serocalculator` therefore floors the raw estimate at 0
(emitting a warning), independent of `floor_to_standard`,
before it can reach $\sqrt{V}$ as a silent `NaN`:

$$V_{\text{raw}}^{+} = \max\paren{0, V_{\text{multiway}}}$$

Setting `floor_to_standard = TRUE` additionally floors the result
at the model-based (non-cluster-robust) variance
$V_{\text{standard}} = H^{-1}$,
the inverse of the observed information at the MLE:

$$V_{\text{final}} = \max\paren{V_{\text{standard}}, V_{\text{raw}}^{+}}$$

If any one-way subset term is unavailable,
for example because its Hessian is non-finite or non-positive,
the inclusion-exclusion sum is itself non-finite.
`serocalculator` warns and returns a missing standard error in that case.
With `floor_to_standard = TRUE` it falls back to $V_{\text{standard}}$
instead, unless $V_{\text{standard}}$ is itself unusable
(the same degenerate Hessian poisons $V_{\text{standard}} = H^{-1}$ too),
in which case it also returns a missing standard error
rather than silently falling back to a negative or infinite "variance".
:::

### Implementation in serocalculator

Users can specify clustering using the `cluster_var` parameter:

```{r, eval=FALSE}
# Single-level clustering (e.g., by household)
est <- est_seroincidence(
  pop_data = data,
  cluster_var = "household_id",
  ...
)

# Multi-level clustering (e.g., schools within districts)
est <- est_seroincidence(
  pop_data = data,
  cluster_var = c("district_id", "school_id"),
  ...
)
```

When cluster-robust standard errors are used, the `summary()` output indicates this with `se_type = "cluster-robust"`.
The multi-way behavior above is itself controlled through `summary()`:

```{r, eval=FALSE}
summary(
  est,
  small_sample = "CR1", # the default; "none" disables the CR1 correction
  floor_to_standard = FALSE, # the default; TRUE additionally floors at V_standard
  debug_cluster = TRUE # prints the V_S/V_raw/V_final decomposition
)
```

::: notes
Internally the optimization is parameterized by $\log\lambda$ rather
than $\lambda$, which keeps the estimate positive, so the sandwich
above is evaluated on that scale and `summary()` converts the result
back with the delta method
($\hSEf{\hat\lambda} = \hat\lambda \cdot \hSEf{\log\hat\lambda}$).
:::

### Effect on results

* **Point estimates** (incidence rates) are unchanged: `cluster_var` affects only the variance, not the likelihood being maximized
* **Standard errors** usually increase, reflecting within-cluster correlation
* **Confidence intervals** widen correspondingly, reflecting the reduced effective sample size
