In many survey designs, observations are clustered (e.g., multiple individuals from the same household, school, or geographic area). Observations within the same cluster are often more similar to each other than to observations from different clusters, violating the independence assumption of standard maximum likelihood estimation.

### Why clustering matters

When observations are clustered:

* Individuals within the same cluster share common exposures or characteristics
* Standard errors that ignore clustering will be **too small** (anti-conservative)
* Confidence intervals will be **too narrow**
* p-values will be **too optimistic**

### Cluster-robust variance estimation

To account for within-cluster correlation, `serocalculator` implements the **sandwich estimator** (also known as the Huber-White robust variance estimator):

$$V_{\text{robust}} = H^{-1} B H^{-1}$$

where:

* $V_{\text{robust}}$ is the cluster-robust variance-covariance matrix for the parameter estimates
* $H$ is the Hessian matrix of the *negative* log-likelihood
  (the observed information),
  evaluated at the MLE $\hl$.
  `serocalculator` obtains it from `stats::nlm()`'s own `hessian` output.
* $B$ is the "meat" of the sandwich, calculated from cluster-level score contributions:

$$B = \sum_{c=1}^C U_c U_c^T$$

where:

* $C$ is the total number of clusters in the sample
* $U_c = \sum_{i \in c} \nabla_\lambda \log p(Y_i | \lambda)$ is the score contribution (gradient of log-likelihood with respect to $\lambda$) from all observations in cluster $c$
* $\nabla_\lambda$ denotes the gradient operator (vector of partial derivatives with respect to the parameter $\lambda$)

### Multi-way clustering

When `cluster_var` names more than one variable,
`serocalculator` does not collapse them into a single interaction.
Instead it applies the inclusion-exclusion identity of
Cameron, Gelbach, and Miller (2011) [@Cameron_2011]
for multi-way cluster-robust variance estimation.

Let $\mathcal{K} = \{k_1, \dots, k_p\}$ denote the cluster variables.
For a non-empty subset $S \subseteq \mathcal{K}$,
let $V_S$ be the one-way sandwich variance $V_{\text{robust}}$ defined above,
computed by clustering on the interaction of the variables in $S$.
The multi-way variance is the signed sum over every such subset:

$$V_{\text{multiway}} = \sum_{\emptyset \neq S \subseteq \mathcal{K}} (-1)^{|S| + 1} V_S$$

For two cluster variables ($p = 2$),
this reduces to the familiar three-term formula:

$$V_{\text{multiway}} = V_1 + V_2 - V_{1 \cap 2}$$

where $V_1$ and $V_2$ cluster on each variable separately,
and $V_{1 \cap 2}$ clusters on their interaction,
the most granular grouping.
Computing every subset costs $2^p - 1$ one-way variance evaluations,
so this is intended for the small $p$ (2-3)
typical of nested survey designs.

### Small-sample correction

`serocalculator` approximates each cluster's score contribution $U_c$
by a forward finite difference of the cluster's log-likelihood,
$\ell_c(\lambda) = \sum_{i \in c} \log p(Y_i \mid \lambda)$:

$$U_c \approx \frac{\ell_c(\hl + \epsilon) - \ell_c(\hl)}{\epsilon}, \quad \epsilon = 10^{-6}$$

When `small_sample = "CR1"` (the default),
each one-way subset variance $V_S$ is inflated by the usual
cluster degrees-of-freedom correction:

$$V_S^{\text{CR1}} = \frac{G_S}{G_S - 1} \, V_S$$

where $G_S$ is the number of unique clusters in subset $S$'s grouping.

### Safeguards against invalid variance

The inclusion-exclusion sum $V_{\text{multiway}}$
is not guaranteed to be non-negative.
This is a documented property of the Cameron-Gelbach-Miller estimator,
not a bug.
`serocalculator` therefore floors the raw estimate at 0
(emitting a warning), independent of `floor_to_standard`,
before it can reach $\sqrt{V}$ as a silent `NaN`:

$$V_{\text{raw}}^{+} = \max\paren{0, V_{\text{multiway}}}$$

Setting `floor_to_standard = TRUE` additionally floors the result
at the model-based (non-cluster-robust) variance
$V_{\text{standard}} = H^{-1}$,
the inverse of the observed information at the MLE:

$$V_{\text{final}} = \max\paren{V_{\text{standard}}, V_{\text{raw}}^{+}}$$

If any one-way subset term is unavailable,
for example because its Hessian is non-finite or non-positive,
the inclusion-exclusion sum is itself non-finite.
`serocalculator` warns and returns a missing standard error in that case.
With `floor_to_standard = TRUE` it falls back to $V_{\text{standard}}$
instead, unless $V_{\text{standard}}$ is itself unusable
(the same degenerate Hessian poisons $V_{\text{standard}} = H^{-1}$ too),
in which case it also returns a missing standard error
rather than silently falling back to a negative or infinite "variance".

### Implementation in serocalculator

Users can specify clustering using the `cluster_var` parameter:

```{r, eval=FALSE}
# Single-level clustering (e.g., by household)
est <- est_seroincidence(
  pop_data = data,
  cluster_var = "household_id",
  ...
)

# Multi-level clustering (e.g., schools within districts)
est <- est_seroincidence(
  pop_data = data,
  cluster_var = c("district_id", "school_id"),
  ...
)
```

When cluster-robust standard errors are used, the `summary()` output indicates this with `se_type = "cluster-robust"`.
The multi-way behavior above is itself controlled through `summary()`:

```{r, eval=FALSE}
summary(
  est,
  small_sample = "CR1", # the default; "none" disables the CR1 correction
  floor_to_standard = FALSE, # the default; TRUE additionally floors at V_standard
  debug_cluster = TRUE # prints the V_S/V_raw/V_final decomposition
)
```

### Effect on results

* **Point estimates** (incidence rates) remain unchanged
* **Standard errors** often increase to reflect within-cluster correlation
* **Confidence intervals** appropriately widen to account for reduced effective sample size
