The estimates above combine HlyE IgA and IgG with the default
`method = "composite"`. @sec-multiple-biomarkers introduced the alternative,
`method = "joint"`, which integrates over a shared latent infection time
per person. The simulated clusters are the place to compare them, because
`sim_pop_data()` draws **one** infection history per person and reads every
biomarker off it: the shared-latent-time structure that the joint likelihood
assumes is exactly how these data were made.

```{r}
#| label: est-multi-joint
#| warning: false
ests_joint <- est_seroincidence_by_cached(
  pop_data = sim_df,
  sr_params = sim_curve_params,
  noise_params = sim_noise_params,
  strata = c("sample_size", "lambda.sim", "cluster"),
  curve_strata_varnames = NULL,
  noise_strata_varnames = NULL,
  antigen_isos = antibodies,
  build_graph = FALSE,
  method = "joint",
  cache_path = sim_cache,
  cache_id = "ests_joint"
)

ests_joint_summary <- ests_joint |> summary()
```

{{< slidebreak >}}

```{r}
#| label: tbl-joint-vs-composite-sim
#| tbl-cap: >
#|   Bias, empirical standard error, mean estimated standard error and
#|   95% confidence-interval coverage of the composite and joint estimators,
#|   by true incidence rate, across the simulated clusters.
sim_comparison <- bind_rows(
  composite = ests_summary |> analyze_sims(),
  joint = ests_joint_summary |> analyze_sims(),
  .id = "method"
) |>
  arrange(lambda.sim, method) |>
  select(
    lambda.sim, method, Bias, Empirical_SE, Mean_Est_SE, CI_Coverage
  )

sim_comparison
```

```{r}
#| label: fig-joint-vs-composite-sim
#| fig-cap: >
#|   Estimated incidence rates (points, with 95% confidence intervals) for
#|   each simulated cluster, against the true rate, under the composite and
#|   the joint likelihood. The red line is the identity.
bind_rows(
  composite = ests_summary,
  joint = ests_joint_summary,
  .id = "method"
) |>
  ggplot() +
  aes(x = lambda.sim, y = incidence.rate, color = method) +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_pointrange(
    aes(ymin = CI.lwr, ymax = CI.upr),
    position = position_dodge(width = .1),
    size = .2
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  labs(
    x = "true incidence rate (per person-year)",
    y = "estimated incidence rate",
    color = "likelihood"
  )
```

::: notes
The composite estimator over-estimates, increasingly so at higher incidence
(as @sec-recover already showed). The joint estimator **under**-estimates,
by more, and its confidence intervals cover the truth far less often. Its
empirical standard error is the smaller of the two, which is the efficiency
gain the shared latent time was expected to bring, but here that gain is
swamped by bias.
:::

{{< slidebreak >}}

The reason is visible in a single high-incidence cluster:

```{r}
#| label: joint-dropped-subjects
sim_cluster <- sim_df |>
  filter(lambda.sim == max(sim_lambdas), cluster == 1)

est_joint_one <- est_seroincidence(
  pop_data = sim_cluster,
  sr_params = sim_curve_params,
  noise_params = sim_noise_params,
  antigen_isos = antibodies,
  method = "joint"
)
```

A person contributes nothing to the joint likelihood when **no** shared
infection time is compatible with all of their readings under any posterior
draw, and this cluster has many such people. Shortly after infection,
antibody levels are in the hundreds or thousands, while the simulated
biological noise is additive on a scale of 0.5. The simulator's response
curve also differs slightly from the likelihood's --- it has a rise phase,
and a baseline carried over from earlier infections --- so one person's IgA
and IgG readings each point to a slightly different time since infection.
Each biomarker on its own is still explained well (the composite method drops
nobody), but no single $t$ reproduces both to within half a unit, so the
joint likelihood discards exactly the people with the most recent
infections, and the estimate falls. The higher the incidence, the more such
people there are.

::: notes
That mechanism is not specific to this simulator. Any mismatch between the
assumed response curve and the real one plays the same role, and so does a
noise model that is narrower than the real disagreement between a person's
biomarkers. `est_seroincidence()` warns whenever a joint fit drops anyone,
and reports how many. The real SEES data, whose noise model includes a
multiplicative (measurement) component, drop very few
(@tbl-joint-vs-composite-sees).

The joint estimator uses the fact that a person has one infection history,
which is more information --- but it also relies on that fact being
reflected in the data *exactly as modeled*. The composite estimator, which
lets each biomarker find its own infection time, is the more robust of the
two, which is one reason it remains the default. Neither is unbiased here:
the simulator's response curve is not the likelihood's, and both estimators
inherit that.
:::
