A serosurvey usually measures more than one antigen-isotype per participant.
The model treats biomarkers as **conditionally independent given the
time since infection**, so their response densities multiply *inside* each
person's integral over $t$:

$$
\dens(Y_1=y_1, Y_2=y_2) =
\int_t \dens(Y_1=y_1|T=t)\,\dens(Y_2=y_2|T=t)\,\dens_\lambda(T=t)\, dt
$$

Note that this is **not** the product of the two single-biomarker
likelihoods: the shared latent infection time $t$ couples them.

`serocalculator` offers two ways of combining biomarkers, selected by the
`method` argument of `log_likelihood()`, `est_seroincidence()` and
`est_seroincidence_by()`:

* `method = "composite"` (the default) evaluates each biomarker's
  likelihood separately and **sums the log-likelihoods**. That is the
  product of the marginals, which treats the biomarkers as independent
  *unconditionally*, each with its own latent infection time. In the
  composite-likelihood literature this is an *independence likelihood*
  [@varin2011overview]: its point estimate is consistent, but its naive
  standard error assumes the biomarker readings are independent.
* `method = "joint"` evaluates the formula above: one integral per person
  over a **shared** $t$, with the per-biomarker densities multiplied inside
  it. This needs the data to identify people (`ids_varname()`), so that
  each person's readings can be paired, and it pairs the posterior draws of
  the curve parameters across biomarkers too, so that a person's biomarkers
  are evaluated under one coherent draw.

`log_likelihood()` evaluates the log-likelihood at a given $\lambda$, which
lets us see the difference numerically. First, the default:

```{r}
#| label: loglik-multi-biomarker
xs_pk <- xs_data |> filter(Country == "Pakistan")
noise_pk <- noise |> filter(Country == "Pakistan")

ll_iga <- log_likelihood(
  pop_data = xs_pk, curve_params = curves, noise_params = noise_pk,
  antigen_isos = "HlyE_IgA", lambda = 0.1
)

ll_igg <- log_likelihood(
  pop_data = xs_pk, curve_params = curves, noise_params = noise_pk,
  antigen_isos = "HlyE_IgG", lambda = 0.1
)

ll_composite <- log_likelihood(
  pop_data = xs_pk, curve_params = curves, noise_params = noise_pk,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"), lambda = 0.1
)

c(IgA = ll_iga, IgG = ll_igg, sum = ll_iga + ll_igg, composite = ll_composite)
```

The two-biomarker value is **exactly** the sum of the separate ones, as
summing log-likelihoods must give. Under the shared-latent-time model it is
not:

```{r}
#| label: loglik-multi-biomarker-joint
ll_joint <- log_likelihood(
  pop_data = xs_pk, curve_params = curves, noise_params = noise_pk,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"), lambda = 0.1,
  method = "joint"
)

c(sum = ll_iga + ll_igg, joint = ll_joint)
```

::: notes
`R/log_likelihood.R` loops over `antigen_isos`, calls `f_dev()` once per
biomarker, and adds the results; `method = "joint"` instead hands all of a
person's readings to `f_dev_joint()`, which integrates once over $t$ in C
(`src/serocalc_joint.c`). With a single biomarker the two agree, up to
numerical integration.

(@sec-multivariate describes work in progress to relax conditional
independence *given* $t$; `method = "joint"` is the model that work
generalizes, and `method = "composite"` is a different, stronger,
independence assumption.)
:::

{{< slidebreak >}}

The two likelihoods call for different standard errors. The joint
likelihood is a genuine likelihood, so the Hessian-based standard error of
@sec-standard-errors applies to it directly. The composite likelihood's
Hessian understates the variance whenever one person's biomarker readings
are correlated (they respond to the same infection), and the fix is the
cluster-robust estimator of @sec-cluster-robust with the *person* as the
cluster, `cluster_var = ids_varname(pop_data)`. On the Pakistan data, the
two agree closely once the composite standard error is corrected:

```{r}
#| label: tbl-joint-vs-composite-sees
#| tbl-cap: >
#|   Incidence estimates for the Pakistan SEES data, combining HlyE IgA and
#|   IgG under the composite likelihood (with the naive and the
#|   person-clustered standard error) and under the joint likelihood.
fit_pk <- function(...) {
  est_seroincidence(
    pop_data = xs_pk,
    sr_params = curves,
    noise_params = noise_pk,
    antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
    ...
  ) |>
    summary(verbose = FALSE) |>
    select(incidence.rate, SE, CI.lwr, CI.upr, log.lik)
}

bind_rows(
  "composite, naive SE" = fit_pk(),
  "composite, person-clustered SE" = fit_pk(cluster_var = "id"),
  "joint" = fit_pk(method = "joint"),
  .id = "estimator"
)
```

::: notes
The joint estimate is a little higher and its log-likelihood is much higher,
but the latter is not a like-for-like comparison: the two methods evaluate
different functions of the data. Which one is *better* has to be settled on
data where the truth is known; @sec-joint-vs-composite-sim does that on the
simulated clusters, and the answer is not simply "the joint one".
:::

{{< slidebreak >}}

Using both biomarkers concentrates the log-likelihood around its maximum,
and so shrinks the standard error, relative to either biomarker alone
(@fig-loglik):

```{r}
#| fig-cap: >
#|   Log-likelihood curves for HlyE IgA alone, HlyE IgG alone, and both
#|   biomarkers combined (composite likelihood).
#| label: fig-loglik
lik_HlyE_IgA <- graph_loglik( # nolint: object_name_linter.
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = "HlyE_IgA",
  log_x = TRUE
)

lik_HlyE_IgG <- graph_loglik( # nolint: object_name_linter.
  previous_plot = lik_HlyE_IgA,
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = "HlyE_IgG",
  log_x = TRUE
)

lik_both <- graph_loglik(
  previous_plot = lik_HlyE_IgG,
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
  log_x = TRUE
)

print(lik_both)
```
