A serosurvey usually measures more than one antigen-isotype per participant.
The current model treats biomarkers as **conditionally independent given the
time since infection**, so their response densities multiply *inside* each
person's integral over $t$:

$$
\dens(Y_1=y_1, Y_2=y_2) =
\int_t \dens(Y_1=y_1|T=t)\,\dens(Y_2=y_2|T=t)\,\dens_\lambda(T=t)\, dt
$$

Note that this is **not** the product of the two single-biomarker
likelihoods: the shared latent infection time $t$ couples them.

`log_likelihood()` evaluates the log-likelihood at a given $\lambda$, which
lets us check that claim numerically:

```{r}
#| label: loglik-multi-biomarker
ll_iga <- log_likelihood(
  pop_data = xs_data, curve_params = curves, noise_params = noise,
  antigen_isos = "HlyE_IgA", lambda = 0.1
)

ll_igg <- log_likelihood(
  pop_data = xs_data, curve_params = curves, noise_params = noise,
  antigen_isos = "HlyE_IgG", lambda = 0.1
)

ll_both <- log_likelihood(
  pop_data = xs_data, curve_params = curves, noise_params = noise,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"), lambda = 0.1
)

c(IgA = ll_iga, IgG = ll_igg, sum = ll_iga + ll_igg, joint = ll_both)
```

The two-biomarker value is **exactly** the sum of the separate ones. That is
worth pausing on, because the formula above says it should not be: one
integral over a shared $t$ does not factor into two separate integrals.

What it tells us is that `log_likelihood()` currently combines biomarkers by
summing their individual log-likelihoods, which is the stronger assumption
that the biomarkers are independent *unconditionally*, each with its own
latent infection time. `est_seroincidence()` optimizes through the same code
path, so this is the estimation model rather than a quirk of this one
function.

Whether the formula or the implementation should change is
[an open question](https://github.com/UCD-SERG/serocalculator/issues/637).
Until it is settled, read a multi-biomarker standard error as assuming
independence across biomarkers.

::: notes
`R/log_likelihood.R` loops over `antigen_isos`, calls `f_dev()` once per
biomarker, and adds the results, with a source comment naming the
independence assumption. Summing log-likelihoods is the log of the product of
the marginals.

(@sec-multivariate describes work in progress to relax conditional
independence *given* $t$; the question above is whether the current code
implements even that much.)
:::

Using both biomarkers concentrates the log-likelihood around its maximum,
and so shrinks the standard error, relative to either biomarker alone
(@fig-loglik):

```{r}
#| fig-cap: >
#|   Log-likelihood curves for HlyE IgA alone, HlyE IgG alone, and both
#|   biomarkers combined.
#| label: fig-loglik
lik_HlyE_IgA <- graph_loglik( # nolint: object_name_linter.
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = "HlyE_IgA",
  log_x = TRUE
)

lik_HlyE_IgG <- graph_loglik( # nolint: object_name_linter.
  previous_plot = lik_HlyE_IgA,
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = "HlyE_IgG",
  log_x = TRUE
)

lik_both <- graph_loglik(
  previous_plot = lik_HlyE_IgG,
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
  log_x = TRUE
)

print(lik_both)
```
