The single survey above recovered a rate close to the one that generated it.
To check that this was not luck, we simulate many clusters across a range of
**known** incidence rates and ask whether the estimates track them.

```{r}
#| label: sim-multi
sim_lambdas <- c(.05, .1, .15, .2, .5, .8) # true incidence rates
sim_nclus <- 20 # clusters per rate

sim_df <- sim_pop_data_multi_cached(
  curve_params = sim_curve_params,
  lambdas = sim_lambdas,
  nclus = sim_nclus,
  sample_sizes = sim_n,
  age_range = sim_lifespan,
  antigen_isos = antibodies,
  renew_params = FALSE,
  add_noise = TRUE,
  noise_limits = sim_noise_limits,
  format = "long",
  cache_path = sim_cache,
  cache_id = "sim_df"
)
```

::: notes
That is `r length(sim_lambdas)` rates times `r sim_nclus` clusters times
`r sim_n` participants --- `r length(sim_lambdas) * sim_nclus * sim_n`
simulated individuals.

Simulating and estimating on that many people is by far the most expensive
work in this document, and it renders to three formats, so the work would
otherwise be repeated three times per build.
`sim_pop_data_multi_cached()` and `est_seroincidence_by_cached()` write their
results to a committed cache directory keyed on their arguments, so a build
reuses them and only recomputes when the simulation settings actually change.
:::

{{< slidebreak >}}

```{r}
#| label: fig-sim-multi-distribution
#| fig-cap: >
#|   Simulated antibody responses by cluster, faceted by antigen-isotype and
#|   by the true incidence rate used to generate each cluster.
#| fig-height: 5
sim_df |>
  ggplot() +
  aes(x = as.factor(cluster), y = value) +
  geom_beeswarm(size = .2, alpha = .3, aes(color = antigen_iso)) +
  geom_boxplot(outlier.colour = NA, fill = NA) +
  scale_y_log10() +
  facet_wrap(~ antigen_iso + lambda.sim, nrow = 2) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  labs(x = "cluster")
```
