---
title: "Generate a simulated cross-sectional sample and estimate seroincidence"
subtitle: "Enteric Fever using HlyE IgG and/or HlyE IgA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulate_xsectionalData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


 This script generates simulated seroresponses for incident infections
 as a Poisson process with frequency `lambda'. Responses are generated for
 antibodies given in `abs' (any subset of 1:n(), names in `antibodies').

 Age range of the simulated cross-sectional record is `lifespan'.

 The size of the sample is `nrep'.

 Each individual is simulated separately, but different antibodies are
 modelled jointly (see "sim-graph.r").

 Longitudinal parameters are calculated for an age: `age.fx' (fixed age).
 However, when `age.fx' is set to NA then the age at infection is used.

 The boolean `renew.params' determines whether each infection uses a
 new set of longitudinal parameters, sampled at random from the
 posterior predictive output of  the longitudinal model. If set to FALSE
 a parameter set is chosen at birth and kept, but:
 1. the baseline antibody levels (y0) are updated with the simulated level
    (just) prior to infection, and
 2. when is.na(age.fx) then the selected parameter sample is updated for the
    age when infection occurs.

 There is also a variable `n.mc': when n.mc==0 then a random MC
 sample is chosen out of the posterior set (1:4000). When `n.mc' is
 given a value in 1:4000 then the chosen number is fixed and reused
 in any subsequent infection. This is for diagnostic purposes.


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(serocalculator)
library(tidyverse)


#load in longitudinal parameters, these are modelled from all SEES cases across all ages and countries 
dmcmc <- getAdditionalData(fileURL = "https://osf.io/download/rtw5k")  
  


mcmcL <- dmcmc %>%
  mutate(y1 = y1-y0) %>%
  mutate(r = r-1) %>%
  tidyr::pivot_longer(
      names_to = "parameter",
      cols = c("y0", "y1", "t1", "alpha", "r") ) |>
      mutate(parameter = factor(parameter, levels = unique(parameter))) %>%
  select(-Country, -ageCat) %>%
  mutate(value = log(value)) %>%
  filter(antigen_iso == "HlyE_IgG" | antigen_iso == "HlyE_IgA") %>%
  droplevels()
  



predpar <- df_to_array(mcmcL, dim_var_names = c("antigen_iso", "parameter"))
 

```


## Simulate cross-sectional data
```{r}
#set seed to reproduce results
set.seed(54321)

ab.nm <- rbind(c("HlyE","IgA"),c("HlyE","IgG"));
antibodies <- c();
for(k.test in 1:nrow(ab.nm))
  antibodies <- c(antibodies,paste(ab.nm[k.test,1],ab.nm[k.test,2]));

antibodies <- levels(mcmcL$antigen_iso)

abs <-  c(1,2);
age.fx <- NA; # age used for longitudinal model (fixed for now)
day2yr <- 365.25;
lambda <- 0.2; # incidence (1/yr)
npar <- 5; # y0, y1, t1, alpha, shape

lifespan <- c(0,20); # range covered in simulations
nrep <- 100; # cross-sectional sample size

dlims <- rbind(c(1e-2,5e-1),c(1e-2,5e-1)); # noise distribution
blims <- dlims; # baseline distribution


#rename function to sim.cs
csdata0 <- makecs(lambda/day2yr,
                  n.smpl=nrep,
                  age.rng=lifespan*day2yr,
                  age.fx=age.fx,
                  ablist=abs,n.mc=0,
                  renew.params=TRUE,add.noise=TRUE);

csdata0[,1] <-  csdata0[,1]/day2yr;


```


## Estimate seroincidence
```{r}

cond <- data.frame(
  antigen_iso = c("HlyE_IgG", "HlyE_IgA"),
  nu = c(0.5, 0.5),  # Biologic noise (nu)
  eps = c(0.2, 0.2),                             # M noise (eps)
  y.low = c(1, 1),                            # low cutoff (llod)
  y.high = c(5e6, 5e6))                      # high cutoff (y.high)
                          

csdataL <- as.data.frame(csdata0) %>%
  pivot_longer(cols= c("HlyE_IgG", "HlyE_IgA"), names_to = "antigen_iso", values_to = "value")

est.incidence(csdataL %>%
                   rename(y = value,
                          a = age), 
              dmcmc, cond, antigen_isos = c("HlyE_IgA"))


```

