---
title: "Enteric Fever Seroincidence Tutorial"
author: "UC Davis Seroepidemiology Research Group (SERG)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Typhoid Seroincidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
toc: true

---
# Introduction

Antibody levels measured in a cross–sectional population sample can be translated into an estimate of the frequency with which seroconversions (infections) occur in the sampled population. In simple 
terms: the presence of many high antibody concentrations indicates that many subjects likely experienced infection recently, while mostly low concentrations indicate a low frequency of infections in the sampled population. 

In order to thus interpret the measured cross-sectional antibody concentrations, characteristics of the time course of the serum antibody response must be known. For any antibody detected in the cross–sectional sample, a quantitative description of its dynamics following infection 
(seroconversion to an elevated concentration, followed by a gradual decrease) must be available. In published studies, this information on the time course of the serum antibody response has been obtained from follow–up data in subjects who had a symptomatic episode following infection. The 
onset of symptoms thus provided a proxy for the time that infection occurred. Care must be taken that longitudinal and cross–sectional measurements of antibody concentrations are calibrated on the 
same scale (or, preferably, expressed in identical units).

The time course of the serum antibody response to infection is assumed generic: anyone infected by the same pathogen is expected to produce a response similar to those seen in the symptomatic cohort used for the follow–up study. Thus, the strategy is to set up a longitudinal model from a follow–up study once, and use the thus obtained information to analyze many different cross–sectional data sets.

Note that it must be assumed that such variation in the time course of antibody concentrations among individuals as observed in the longitudinal cohort is the same as in the population where the cross-sectional sample was taken, where this variation cannot be observed.

Analysis of longitudinal data and estimation of the required parameters is separated from the estimation of seroincidences, because (1) there are few longitudinal data sets available and they are difficult (and expensive) to obtain, (2) such longitudinal models may be considered *generic*, presumably describing a physiological phenomenon that is similiar in any infected individual, and (3) setting up and running the longitudinal model requires additional skills, not necessary for use of the sero–incidence calculator script.

# The seroincidence estimator

The package **serocalculator** was designed to calculate incidence of seroconversion, by using the longitudinal seroresponse characteristics. The distribution of serum antibody concentrations in a cross–sectional population sample is calculated, as a function of the longitudinal seroresponse 
and the frequency of seroconversion (or seroincidence). Given the seroresponse, this marginal distribution of antibody concentrations can be fitted to the cross-sectional data, by adjusting the seroconversion frequency, thus providing a means to estimate the seroconversion frequency.

# Vignette

This vignette provides users with an example analysis using the **serocalculator** package by reproducing an analysis in [@Aiemjoy_2022_Lancet]. Users will determine the seroincidence of enteric fever in a cross-sectional serosurvey from SEES, a serologic and environmental surveillance study conducted in Asia, plus existing longitudinal antibody dynamics collected from Bangladesh, Ghana, Nepal, and Pakistan. 

Further details on this study can be found here: https://doi.org/10.1016/S2666-5247(22)00114-8. 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Sample Analysis
## Load packages
The first step in conducting this analysis is to load our necessary packages. 

```{r setup, message=FALSE}

#library(devtools)
#install_github("UCD-SERG/serocalculator")
library(serocalculator)
library(tidyverse)
library(ggplot2)
library(readr)


```
## Load data



### Pulling data from Open Science Framework (OSF)

Sample datasets and pathogen-specific longitudinal antibody dynamics for **serocalculator** are housed on Open Science Framework (OSF), a publicly available data repository. The [Serocalculator Repository](https://osf.io/ne8pc/) can be called directly into the **serocalculator** package using the function `getAdditionalData`. 

To import the provided datasets into your R environment, we will need to identify the "id" of the appropriate dataset. 

We demonstrate this below. 

```{r getdata}
getAdditionalData(fileURL = "https://osf.io/download/6553f989874c2e06a54e7d45/")


```


### Load and prepare longitudinal paramter data


### a. Load and prepare longitudinal parameter data
The next step is to load the longitudinal data to set the antibody decay parameters. In this example, these parameters were modeled with Bayesian hierarchical models to fit two-phase power-function decay models to the longitudinal antibody responses among confirmed enteric fever cases. 

These parameters include the following:

  * $y_0$ = baseline  
  * $y_1$ = peak antibody responses 
  * $t_1$ = time to peak 
  * $\alpha$ = decay rate
  * $r$ = decay shape 

We also create two additional variables: the annual decay rate, $\alpha$, which is calculated from the daily decay rate in this example, and $d$, which is the decay shape, or 1-$r$.

Finally, we select only the variables needed for the analysis. 

```{r longdata, echo=FALSE, message=FALSE}
c.hlye.IgG <-                             #Need to change this to pull from osf
  fs::path_package(                          
  "extdata", 
  "dmcmc_hlyeigg_09.30.rds", 
  package = "serocalculator") |> #Load longitudinal parameters dataset
 readRDS()%>%
  mutate(alpha = alpha*365.25, #Create alpha and d 
         d = r-1) %>%
  select(y1, alpha, d) #Select only the variables needed for analysis
```

### Load and prepare cross-sectional data 
The simulated data represents a cross-sectional serosurvey conducted in a representative sample of the general population without regard to disease status. 

In this scenario, we have selected hlye and IgG as our target measures. Users may select different serologic markers depending on what is available. From the original dataset, we rename our variables to *y* and *a*. Finally, we once again limit the dataset to only the variables needed for the analysis. 

``` {r simdata, message=FALSE}
library(fs) # filesystem utility functions
p.hlye.IgG  <- 
  fs::path_package(
    package = "serocalculator", 
    "extdata/simpophlyeigg.2.csv") %>% #Load  cross-sectional dataset
  read_csv() %>%
  rename( #rename variables
    y = y.smpl,
    a = a.smpl) %>% 
  select(y, a) #Select only the variables needed for analysis
```


### Set conditions for simulated data (Should this section be deleted?)
Next, we must set conditions based on some assumptions about the simulated data. This will differ based on background knowledge of the cross-sectional data. 

The biological noise, $\nu$ ("nu"), represents error from cross-reactivity to other antibodies in addition to those for the target condition. This can artificially inflate quantitative antibody result. Measurement noise, $\varepsilon$ ("epsilon"), represents error from the laboratory testing process. 

``` {r conditions, message=FALSE}
cond.hlye.IgG <- data.frame(
  nu = 1.027239,             # Biological noise
  eps = 0.2,            # Measurement noise
  y.low = 0.0,          # low cutoff
  y.high = 5e4); # high cutoff
```
## 3. Visualize antibody data
Get to know your cross-sectional antibody data by visualizing the distribution of quantitative antibody responses. We have selected HlyE as the antigen of interest, and both IgA and IgG as the isotypes of interest. Here, we will look at the distribution of HlyE IgA and HlyE IgG. 
```{r hist}
#Graph your data as a histogram
hist.hlye.IgG <- ggplot(p.hlye.IgG, aes(x=y)) + geom_histogram(binwidth=5, color = "#000000", fill = "#009999")+  theme_test() + ggtitle("Distribution of Cross-sectional Antibody Responses") 

hist.hlye.IgG
```

Here, we see that our data is highly skewed with the majority of responses on the lower end of our data, but there is a long tail stretching to over 300. Let's get a better look at the distribution by log transforming our *y* response. We will also need to update the bin width. 

```{r loghist}
#Let's log transform your data to better visualize the distribution. We will also need to update the bin width. 
loghist.hlye.IgG <- ggplot(p.hlye.IgG, aes(x=y)) + geom_histogram(binwidth=0.1, color = "#000000", fill = "#009999") +scale_x_log10() +ggtitle("Distribution of Cross-sectional Antibody Responses (Log transformed)") + xlab("Log(y)")

loghist.hlye.IgG
```

Once log transformed, our data looks much more normally distributed. In most cases, log transformation will be the best way to visualize serologic data. 

## Estimate Seroincidence 
Finally, we are ready to begin seroincidence estimation. We will conduct two separate analyses using two distinct functions, `est.incidence` and `est.incidence.by`, to calculate the overall seroincidence and the age-specific incidence, respectively.

### Overall Seroincidence
We define our starting value as 0.5, which will also define our initial estimate for the force of infection (FOI, $\lambda$ ("lambda")). Then we log transform $\lambda$ and set up maximum and minimum values for the confidence interval. 


```{r seroinc}
start <- .05 #Set starting value

lambda = start #initial estimate: starting value
log.lambda = log(lambda)
log.lmin=log(lambda/10)
log.lmax=log(10*lambda) 


objfunc <- function(llam){
  return(res <- fdev(llam, p.hlye.IgG, c.hlye.IgG, cond.hlye.IgG))
}

fit <- nlm(objfunc,log.lambda,
           hessian=TRUE,print.level=0,stepmax=(log.lmax-log.lmin)/4)

#Calculate lambda, lower, upper, LF min
log.lambda.est <- c(exp(fit$estimate),
                    exp(fit$estimate + qnorm(c(0.025))*sqrt(1/fit$hessian)),
                    exp(fit$estimate + qnorm(c(0.975))*sqrt(1/fit$hessian)),
                    fit$minimum)

#Print the final results
log.lambda.est
```

### Age-Specific Seroincidence

## Conclusions
In our simulated data, we found that the estimated seroincidence of enteric fever is 0.20 (95% CI: 0.18, 0.23). 


## References
