---
title: "Methodology"
vignette: >
  %\VignetteIndexEntry{Methodology}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

{{< include articles/_macros.qmd >}}

# Overview

## Defining incidence

The **incidence rate** of a disease over a **specific time period** is 
the rate at which individuals in a population are *acquiring* the disease 
during that time period [@Noordzij2010diseasemeasures].

**Example:**
if there are **10 new cases** of typhoid 
in a population of **1000 persons** 
during a **one month** time period, 
then the incidence rate for that time period is
**10 new cases per 1000 persons per month**.

## Mathematical definition of incidence {visibility="hidden"}

More mathematically, 
the incidence rate *at a given time point* is 
the *derivative* (i.e., the current rate of change over time) 
of the expected cumulative count of infections per person at risk, 
at that time:

$$\frac{d}{dt} \mathbb{E}\left[\frac{C(t)}{n} \mid N(t) =n\right]$$

where $C(t)$ is the cumulative total number of infections in 
the population of interest,
and $N(t)$ is the number of individuals at risk at time $t$.

## Scale of incidence rates {visibility="hidden"}

In both definitions, 
the units for an incidence rate are 
"# new infections per # persons at risk per time duration"; 
for example, "new infections per 1000 persons per year". 

For convenience, 
we can rescale the incidence rate to make it easier to understand; 
for example, we might express incidence as 
"# new infections per 1000 persons per year" 
or "# new infections per 100,000 persons per day", etc.

## Incidence from an individual's perspective

From the perspective of an individual in the population:

* the **incidence rate** (at a given time point ($t$)
is the instantaneous **probability** (density) of 
**becoming infected** at that time point,
**given** that they are **at risk** at that time point.

* That is, the incidence rate is a **hazard** rate.

* Notation: let's use **$\lambda_{t}$** to denote the incidence rate at time $t$.

# Estimating incidence from cross-sectional antibody surveys

## Cross-sectional antibody surveys {.incremental}

::: notes
Typically, it is difficult to estimate changes from a single time point. 
However, we can sometimes make assumptions that allow us to do so. 
In particular, if we assume that the incidence rate is constant over time, 
then we can estimate incidence from a single cross-sectional survey.

We will need two pieces of notation to formalize this process.

:::

::: incremental

* We recruit participants from the population of interest.

* For each survey participant, we measure antibody levels $(Y)$ 
for the disease of interest

* Each participant was **most recently infected** at some time $(T)$
**prior** to when we measured their antibodies.

::: notes
* If a participant has never been infected since birth, then $T$ is undefined.


:::

* $T$ is a **latent, unobserved variable**.

::: notes
* We **don't directly observe $T$**; we **only observe $Y$**, 
which we hope tells us something about $T$ and $\lambda$.

:::

:::

## Modeling assumptions

We **assume** that:

::: incremental

* The incidence rate is approximately **constant** 
**over time** and **across the population** ("**constant and homogenous incidence**")

::: notes

* that is:
$$\lambda_{i,t} = \lambda, \forall i,t$$

(We can analyze subpopulations separately to make homogeneity more plausible.)
:::

* Participants are always at risk of a new infection, 
regardless of how recently they have been infected
("**no lasting immunity**").

:::

::: notes
(For diseases like typhoid, 
the no-immunity assumption may not hold exactly,
but hopefully approximately; 
modeling the effects of re-exposure during an active infection is 
[on our to-do list](https://github.com/UCD-SERG/dcm/issues/11)).
:::

## Time since infection and incidence

Under those assumptions:

::: incremental

* $T$ has an **exponential distribution**:

$$\pdf(T=t) = \lambda \expf{-\lambda t}$$

  * More precisely, the distribution is exponential 
  **truncated by age** at observation ($a$):

$$\pdf(T=t|A=a) = 1_{t \in[0,a]}\lambda \expf{-\lambda t} + 1_{t = \NA} \expf{-\lambda a}$$

* the rate parameter $\lambda$ is the incidence rate

:::

::: notes
This is a time-to-event model, 
looking **backwards in time** from the survey date 
(when the blood sample was collected).

The probability that 
an individual was **last** infected $t$ days ago, $p(T=t)$, 
is equal to the probability of being infected at time $t$ 
(i.e., the incidence rate at time $t$, $\lambda$) 
times the probability of not being infected after time $t$, 
which turns out to be $\exp(-\lambda t)$.

The distribution of $T$ is truncated by the patient's birth date;
the probability that they have never been infected is
$\expf{-\lambda a}$, 
where $a$ is the patient's age at the time of the survey.
:::

## Likelihood of latent infection times {.smaller}

{{< include articles/_sec-latent-likelihood.qmd >}}

## Example log-likelihood curves

::: notes
Here's what that would look like:
:::

```{r}
#| fig-cap: "Example log-likelihood curves"
#| label: fig-ex-lik-curves

library(serocalculator)
library(dplyr)
# Import longitudinal antibody parameters from OSF
curves <-
  "https://osf.io/download/rtw5k/" %>%
  load_sr_params() %>%
  filter(iter < 50)

# Import cross-sectional data from OSF and rename required variables:
xs_data <-
  "https://osf.io/download//n6cp3/" %>%
  load_pop_data()

noise <- url("https://osf.io/download//hqy4v/") %>% readRDS()
```

```{r}
#| fig-cap: "Example log(likelihood) curves"
#| label: fig-loglik
lik_HlyE_IgA <- graph_loglik(
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = "HlyE_IgA",
  log_x = TRUE
)

lik_HlyE_IgG <- graph_loglik(
  previous_plot = lik_HlyE_IgA,
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = "HlyE_IgG",
  log_x = TRUE
)

lik_both <- graph_loglik(
  previous_plot = lik_HlyE_IgG,
  pop_data = xs_data,
  curve_params = curves,
  noise_params = noise,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
  log_x = TRUE
)

print(lik_both)
```

## standard error

The standard error of the estimate is approximately equal to the inverse of the rate of curvature (2nd derivative, aka Hessian) in the log-likelihood function, at the maximum:

more curvature -> likelihood peak is clearer -> smaller standard errors

## Likelihood of observed data

::: notes
Unfortunately, we don't observe infection times $T$; 
we only observe antibody levels ${Y}$. 
So things get a little more complicated.

In short, we are hoping that we can estimate 
$T$ (time since last infection) 
from $Y$ (current antibody levels). 
If we could do that, 
then we could plug in our estimates $\hat t_i$ 
into that likelihood above, 
and estimate $\lambda$ as previously.

We're actually going to do something a little more nuanced; 
instead of just using one value for $\hat t$, 
we are going to consider all possible values of $t$ for each individual.

We need to link the data we actually observed to the incidence rate.

The likelihood of an individual's observed data, 
$\pdf(Y=y)$, 
can be expressed as an integral over 
the joint likelihood of $Y$ and $T$ 
(using the Law of Total Probability):
:::

::: incremental

* $$\pdf(Y=y) = \int_t \pdf(Y=y,T=t)dt$$

:::: notes
Further, we can express the joint probability $p(Y=y,T=t)$ as the product of 
$p(T=t)$ and 
$p(Y=y|T=t)$ the "antibody response curve after infection". 
That is:
::::

* $$\pdf(Y=y,T=t) = \pdf(Y=y|T=t) \pdf(T=t)$$

:::

## Antibody response curves

::: {#fig-decay}

![](fig/fig1a-1.svg)

Antibody response curves, $p(Y=y|T=t)$, for typhoid

:::

## Putting it all together

Substituting $p(Y=y,T=t) = p(Y=y|T=t)P(T=t)$ 
into the previous expression for $p(Y=y)$:

$$
\begin{aligned}
p(Y=y)
&= \int_t p(Y=y|T=t)P(T=t) dt
\end{aligned}
$$

## Composing the likelihood {.smaller}

::: notes
Now, the likelihood of the observed data 
$\vec{y} = (y_1, y_2, ..., y_n)$ is:
:::

$$
\begin{aligned}
\mathcal{L}(\lambda) 
&= \prod_{i=1}^n p(Y=y_i)
\\&= \prod_{i=1}^n \int_t p(Y=y_i|T=t)p_\lambda(T=t)dt\\
\end{aligned}
$$

::: notes
If we know $p(Y=y|T=t)$, then we can maximize $\mathcal{L}(\lambda)$ over $\lambda$ to find the "maximum likelihood estimate" (MLE) of $\lambda$, denoted $\hat\lambda$.
:::

## Finding the MLE numerically

::: notes
The likelihood of $Y$ involves the product of integrals, so the log-likelihood involves the sum of the logs of integrals:
:::

$$
\begin{aligned}
\log \mathcal{L} (\lambda) 
&= \log \prod_{i=1}^n \int_t p(Y=y_i|T=t)p_\lambda(T=t)dt\\
&= \sum_{i=1}^n \log\left\{\int_t p(Y=y_i|T=t)p_\lambda(T=t)dt\right\}\\
\end{aligned}
$$

::: notes

The derivative of this expression doesn't come out cleanly, 
so we will use a *numerical method* 
(specifically, a Newton-type algorithm, implemented by `stats::nlm()`) 
to find the MLE and corresponding standard error.

:::

## Cluster-robust standard errors for clustered sampling designs

{{< include articles/_cluster-robust-se.qmd >}}

{{< include articles/_antibody-response-model.qmd >}}

<!-- {{< include articles/_methods-continued.qmd >}} -->

## References {.smaller .unnumbered}

::: {#refs}
:::
