{{< include macros.qmd >}}

# Modeling the seroresponse kinetics curve

::: notes
Now, we need a model for the antibody response to infection, $\pdf(Y=y|T=t)$.
The current version of the `{serocalculator}` package uses
a two-phase model for
the shape of the seroresponse [@Teunis_2016].
:::

## Model for active infection period {.smaller}

::: notes
The first phase of the model represents the **active infection period**,
and uses a simplified Lotka-Volterra predator-prey model
[@volterra1928variations]
where the pathogen is the prey and the antibodies are the predator:
:::

Notation:

* $x(t)$: Pathogen concentration at time $t$ 
* $y(t)$: Antibody concentration at time $t$ 

Model: 

* $$x'(t) = \alpha x(t) - \beta y(t)$$
* $$y'(t) = \delta y(t)$$

::: notes
With baseline antibody concentration $y(0) = y_{0}$ and
initial pathogen concentration
$x(0) = x_{0}$.
:::

::: notes
Compared to the standard LV model:

* the predation term with the $\beta$ coefficient is missing
the prey concentration $x(t)$ factor;
we assume that the efficiency of predation
doesn't depend on pathogen concentration.

* the differential equation for predator density is missing
the predator death rate term
$-\gamma y(t)$; we assume that as long as there are any pathogens present,
the antibody decay rate is negligible compared to the growth rate.

* the predator growth rate term $\delta y(t)$ is missing
the prey density factor $x(t)$
we assume that as long as there are any pathogens present,
the antibody concentration grows at the same exponential rate.

These omissions were made to simplify the estimation process,
under the assumption that they are negligible
compared to the other terms in the model.

:::

## Model for post-infection antibody decay

* $$b(t) = 0$$

* $$y^{\prime}(t) = -\alpha y(t)^r$$

::: notes
If $r = 1$, this model reduces to exponential decay with decay rate $\alpha$.
:::

## Putting it all together

The serum antibody response $y(t)$ can be written as

$$
y(t) = y_{+}(t) + y_{-}(t)
$$

where

$$
\begin{align}
y_{+}(t) & = y_{0}\text{e}^{\mu t}[0\le t <t_{1}]\\
y_{-}(t) & = y_{1}\left(1+(r-1)y_{1}^{r-1}\alpha(t-t_{1})\right)^{-\frac{1}{r-1}}[t_{1}\le t < \infty]
\end{align}
$$

---

Since the peak level is $y_{1} = y_{0}\text{e}^{\mu t_{1}}$
the growth rate $\mu$ can be written as
$$\mu = \frac{1}{t_{1}}\log\left(\frac{y_{1}}{y_{0}}\right)$$

---

```{r}
cur_ai = "HlyE_IgG"
```

```{r}
#| label: fig-response-graph
#| fig-cap: An example kinetics curve for HlyE IgG

library(serocalculator)
library(dplyr)
# Import longitudinal antibody parameters from OSF
curves <-
  "https://osf.io/download/rtw5k/" %>%
  load_curve_params() %>%
  filter(iter < 50)

curve1 =
  curves %>%
  filter(
    # iter %in% 1:10,
    iter == 5,
         antigen_iso == cur_ai)

library(ggplot2)

curve1 %>%
serocalculator:::plot_curve_params_one_ab(
  log_y = FALSE
) +
  xlim(0, 100) +
  theme_minimal() +
  geom_vline(
    aes(xintercept = curve1$t1,
        col = "t1")
  ) +

  geom_hline(
    aes(yintercept = curve1$y0,
        col = "y0")
  ) +


  geom_hline(
    aes(yintercept = curve1$y1,
        col = "y1")
  ) +
  # geom_point(
  #   data = curve1,
  #   aes(
  #     x = t1,
  #     y = y1,
  #     col = "(t1,y1)"
  #   )
  # ) +
  theme(legend.position = "bottom") +
  labs(col = "")

```


:::: notes
The antibody level at $t=0$ is $y_{0}$;
the rising branch ends at $t = t_{1}$
where the peak antibody level $y_{1}$ is reached.
Any antibody level $y(t) \in (y_{0}, y_{1})$ eventually occurs twice.
::::

---

Antibody decay is different from exponential (log--linear) decay. When the shape parameter $r > 1$,
log concentrations decrease rapidly after infection has terminated,
but decay then slows down and
low antibody concentrations are maintained for a long period.
When $r$ approaches 1, exponential
decay is restored.


## Biological noise

When we measure antibody concentrations in a blood sample, we are essentially counting molecules (using biochemistry).

We might miss some of the antibodies (undercount, false negatives) and we also might incorrectly count some other molecules that aren't actually the ones we are looking for (overcount, false positives, cross-reactivity).

We are more concerned about overcount (cross-reactivity) than undercount. For a given antibody, we can do some analytical work beforehand to estimate the distribution of overcounts, and add that to our model $p(Y=y|T=t)$.

## Measurement noise

There are also some other sources of noise in our bioassays; user differences in pipetting technique, random ELISA plate effects, etc. This noise can cause both overcount and undercount. We can also estimate the magnitude of this noise source, and include it in $p(Y=y|T=t)$.
