# Modeling the seroresponse kinetics curve

::: notes
Now, we need a model for the antibody response to infection, $\dens(Y=y|T=t)$.
The current version of the `{serocalculator}` package uses
a two-phase model for
the shape of the seroresponse [@Teunis_2016].
:::

## Model for active infection period {.smaller}

::: notes
The first phase of the model represents the **active infection period**,
and uses a simplified Lotka-Volterra predator-prey model
[@volterra1928variations]
where the pathogen is the prey and the antibodies are the predator:
:::

Notation:

* $x(t)$: pathogen concentration at time $t$
* $y(t)$: antibody concentration at time $t$
* $\mu_0$: pathogen growth rate; $\beta$: pathogen-inactivation strength; $\mu$: antibody growth rate

Model:

* $$x'(t) = \mu_0\, x(t) - \beta y(t)$$
* $$y'(t) = \mu\, y(t)$$

::: notes
With baseline antibody concentration $y(0) = y_{0}$ and
initial pathogen concentration
$x(0) = x_{0}$.
:::

::: notes
Compared to the standard LV model:

* the predation term with the $\beta$ coefficient is missing
the prey concentration $x(t)$ factor;
we assume that the efficiency of predation
doesn't depend on pathogen concentration.

* the differential equation for predator density is missing
the predator death rate term
$-\gamma y(t)$; we assume that as long as there are any pathogens present,
the antibody decay rate is negligible compared to the growth rate.

* the predator growth rate term $\mu\, y(t)$ is missing
the prey density factor $x(t)$;
we assume that as long as there are any pathogens present,
the antibody concentration grows at the same exponential rate.

These omissions were made to simplify the estimation process,
under the assumption that they are negligible
compared to the other terms in the model.

:::

## Model for post-infection antibody decay

::: notes
Once the immune response clears the infection, the pathogen concentration
reaches zero, so antibody production is no longer stimulated and the
antibody concentration decays.
:::

Once the pathogen is cleared ($x(t) = 0$), the antibody concentration decays:

* $$y^{\prime}(t) = -\alpha y(t)^r$$

::: notes
Antibody decay is different from exponential (log--linear) decay. 
When the shape parameter $r > 1$,
log concentrations decrease rapidly after infection has terminated,
but decay then slows down and
low antibody concentrations are maintained for a long period.
If $r = 1$, 
this model reduces to exponential decay with decay rate $\alpha$.
:::

## Assembling the full response curve

The serum antibody response $y(t)$ can be written as

$$
y(t) = y_{+}(t) + y_{-}(t)
$$

where

$$
\begin{align}
y_{+}(t) & = y_{0}\text{e}^{\mu t}[0\leq t <t_{1}]\\
y_{-}(t) & = y_{1}\left(1+(r-1)y_{1}^{r-1}\alpha(t-t_{1})\right)^{-\frac{1}{r-1}}[t_{1}\le t < \infty]
\end{align}
$$

## Growth rate from peak and baseline {visibility="hidden"}

Since the peak level is $y_{1} = y_{0}\text{e}^{\mu t_{1}}$
the growth rate $\mu$ can be written as
$$\mu = \frac{1}{t_{1}}\log\left(\frac{y_{1}}{y_{0}}\right)$$

## From the full model to the observed curve {.smaller}

::: notes
The antibody curve $y(t)$ that we observe and plot falls out of the
two-phase system as follows:
:::

* **During infection** ($t < t_1$): the antibody equation
$y'(t) = \mu\, y(t)$ gives exponential growth, $y(t) = y_0\, e^{\mu t}$.

* The **pathogen equation sets the peak time** $t_1$: circulating
antibodies inactivate the pathogen (the $-\beta y(t)$ term), driving $x(t)$
to zero at $t_1$. Once the pathogen is cleared, stimulation stops and decay
begins, so the peak antibody level is $y_1 = y_0\, e^{\mu t_1}$.

* **After the peak** ($t \ge t_1$): power-law decay,
$y'(t) = -\alpha y(t)^r$.

::: notes
So the pathogen trajectory $x(t)$ never appears in the final antibody curve
--- it enters only through the peak time $t_1$. From antibody data alone
the pathogen-side quantities ($\mu_0$, $\beta$, and the inoculum $x_0$) are
not separately identifiable [@Teunis_2016], so the curve is summarized by
five parameters: $y_0,\ y_1,\ t_1,\ \alpha,\ r$.
:::

---

```{r}
cur_ai <- "HlyE_IgG"
```

```{r}
#| label: fig-response-graph
#| fig-cap: An example kinetics curve for HlyE IgG

library(serocalculator)
library(dplyr)
# Import longitudinal antibody parameters from OSF
curves <-
  "https://osf.io/download/rtw5k/" |>
  load_sr_params() |>
  filter(iter < 50)

curve1 <-
  curves |>
  filter(
    iter == 5,
    antigen_iso == cur_ai
  )
library(ggplot2)

curve1 |>
  # plot_curve_params_one_ab() picks rows via `iter == seq_len(nrow(.))`, so a
  # single pre-filtered curve must be relabelled iter 1 or no curve is drawn.
  mutate(iter = 1) |>
  serocalculator:::plot_curve_params_one_ab(
    log_y = FALSE
  ) +
  xlim(0, 100) +
  theme_minimal() +
  geom_vline(
    aes(
      xintercept = curve1$t1,
      col = "t1"
    )
  ) +

  geom_hline(
    aes(
      yintercept = curve1$y0,
      col = "y0"
    )
  ) +


  geom_hline(
    aes(
      yintercept = curve1$y1,
      col = "y1"
    )
  ) +
  geom_point(
    data = curve1,
    aes(
      x = t1,
      y = y1,
      col = "(t1,y1)"
    )
  ) +
  theme(legend.position = "bottom") +
  labs(col = "")
```


:::: notes
The antibody level at $t=0$ is $y_{0}$;
the rising branch ends at $t = t_{1}$
where the peak antibody level $y_{1}$ is reached.
Any antibody level $y(t) \in (y_{0}, y_{1})$ eventually occurs twice.
::::

---

::: notes
An interactive Shiny app lets you manipulate the model parameters:
:::

Interactive Shiny app:
<https://ucdserg.shinyapps.io/antibody-kinetics-model-2/>

```{r}
#| echo: false
#| eval: false
#| label: qr-code
library(qrcode)
code <- qr_code("https://ucdserg.shinyapps.io/antibody-kinetics-model-2/")
generate_svg(code, filename = "man/figures/qr_akm2.svg")
```

![QR code for the antibody-kinetics Shiny app](../man/figures/qr_akm2.svg){width="280"}

## Biological noise

::: notes
When we measure antibody concentrations in a blood sample, we are essentially counting molecules (using biochemistry).

We might miss some of the antibodies (undercount, false negatives) and we also might incorrectly count some other molecules that aren't actually the ones we are looking for (overcount, false positives, cross-reactivity).

We are more concerned about overcount (cross-reactivity) than undercount. For a given antibody, we can do some analytical work beforehand to estimate the distribution of overcounts, and add that to our model $p(Y=y|T=t)$.
:::

Notation:

* $y_\text{obs}$: measured serum antibody concentration
* $y_\text{true}$: "true" serum antibody concentration
* $\epsilon_b$: noise due to probe cross-reactivity

Model:

* $y_\text{obs} = y_\text{true} + \epsilon_b$
* $\epsilon_b \sim \text{Unif}(0, \nu)$

$\nu$ needs to be pre-estimated using negative controls,
typically using the 95th percentile of the distribution of
antibody responses to the antigen-isotype
in a population with no exposure.

::: notes
Why the 95th percentile of negative controls, rather than fitting the
full shape of the noise distribution?

[@Teunis_2020] show that a uniform noise model only needs to get the
noise *width* right, not its exact *shape*: comparing $\lambda$
estimates under the true (lognormal) noise distribution against
estimates under a uniform distribution with the same one-sided 95%
range, the two are similar. Getting the width right matters far more
than getting the shape right, so a single width parameter -- rather
than a fully specified distribution -- is sufficient.

That still leaves the question of *how* to estimate the width. The
general cross-sectional sample cannot be used directly: it is a
mixture of never-infected subjects (pure noise) and subjects at
various stages of a real antibody response (signal plus noise), and
separating the two requires already knowing who is infected -- exactly
what the model is trying to estimate. [@Teunis_2020] note this
explicitly: "against a background of ongoing seroresponses, even
reliable verification of the (95%) width of the distribution may be
difficult."

A negative-control panel (subjects confirmed never infected) avoids
this circularity: every observed value in that panel *is* noise, with
no signal mixed in, so its empirical 95th percentile is a direct
estimate of the width the model needs.
:::

## Measurement noise

There are also some other sources of noise in our bioassays; 
user differences in pipetting technique, 
random ELISA plate effects, etc. 
This noise can cause both overcount and undercount. 
We can also estimate the magnitude of this noise source
and include it in $p(Y=y|T=t)$.

Measurement noise, $\varepsilon$ ("epsilon"),
represents measurement error from the laboratory testing process.

Unlike biological noise, measurement noise is *multiplicative*:
the error scales with the true concentration
(equivalently, it is additive on a log scale),
and it has mean zero,
so on average it neither inflates nor deflates the measured concentration.

Notation:

* $y_\text{obs}$: measured serum antibody concentration
* $y_\text{true}$: "true" serum antibody concentration
* $\xi$: relative measurement error
* $\varepsilon$: bound on the relative error

Model:

* $y_\text{obs} = y_\text{true} \cdot (1 + \xi)$
* $\xi \sim \text{Unif}(-\varepsilon, \varepsilon), \quad 0 \le \varepsilon < 1$

Here $\varepsilon$ is the bound on the relative error,
not a coefficient of variation (CV).
Assay precision is often reported as a CV,
the ratio of the standard deviation to the mean for replicates,
ideally measured across plates rather than within the same plate.
Under this uniform model the CV equals $\varepsilon/\sqrt{3}$,
so a measured CV corresponds to $\varepsilon = \sqrt{3}\,\text{CV}$.

::: notes
This follows directly from the variance of a uniform distribution.
For $\xi \sim \text{Unif}(-\varepsilon, \varepsilon)$,

$$
\text{Var}(\xi) = \frac{(2\varepsilon)^2}{12} = \frac{\varepsilon^2}{3}
\quad\Longrightarrow\quad
\text{SD}(\xi) = \frac{\varepsilon}{\sqrt{3}}.
$$

Since $E[\xi] = 0$ by symmetry, for a fixed true concentration
$y_\text{true}$,

$$
E[y_\text{obs}] = y_\text{true} \cdot E[1 + \xi] = y_\text{true},
\qquad
\text{SD}(y_\text{obs}) = y_\text{true} \cdot \text{SD}(\xi)
  = y_\text{true} \cdot \frac{\varepsilon}{\sqrt{3}},
$$

so

$$
\text{CV} = \frac{\text{SD}(y_\text{obs})}{E[y_\text{obs}]}
  = \frac{\varepsilon}{\sqrt{3}}.
$$

The true concentration $y_\text{true}$ cancels, so the CV depends only
on $\varepsilon$, not on the concentration being measured -- as
expected for a *relative* (multiplicative) error model.
:::

## Combined biological and measurement noise

In practice both noise sources are usually present at once.
The two are applied in sequence [@Teunis_2020]:
biological noise is added to the true concentration,
and measurement noise then scales the result.

Model:

* $y_\text{obs} = (y_\text{true} + \epsilon_b)(1 + \xi)$
* $\epsilon_b \sim \text{Unif}(0, \nu)$
* $\xi \sim \text{Unif}(-\varepsilon, \varepsilon)$

This is the model `serocalculator` uses
whenever a `noise_params` row has both $\nu > 0$ and $\varepsilon > 0$;
setting either width to zero recovers the corresponding single-source model above.

## Noise and never-infected subjects

A subject who has never been infected has
true antibody concentration $y_\text{true} = 0$;
the probability of never having been infected by age $a$
is $\expf{-\lambda a}$.
The two noise sources treat such a subject very differently [@Teunis_2020]:

* **Biological noise is additive**,
so a never-infected subject is measured as
$y_\text{obs} = 0 + \epsilon_b \sim \text{Unif}(0, \nu)$.
Cross-reactivity can register a positive signal
even when there was no true response,
so never-infected subjects still contribute
a spread-out distribution of small positive values,
not a single spike at zero.

* **Measurement noise is multiplicative**,
so a never-infected subject is measured as
$y_\text{obs} = 0 \cdot (1 + \xi) = 0$.
A relative error cannot move a true zero,
so measurement noise alone leaves these subjects exactly at zero.

So the probability mass at $y = 0$ from never-infected subjects
(with weight $\expf{-\lambda a}$)
survives measurement noise
but is smeared into small positive values by biological noise.
This also explains why the biological-noise width $\nu$
can be estimated from a known unexposed population:
those subjects are essentially all never-infected,
so their measured antibody levels are, to good approximation,
draws from the biological-noise distribution itself.

**Under both noise sources together**, the never-infected contribution
is neither of these two single-source shapes.
Substituting $y_\text{true} = 0$ into the combined model gives

$$
y_\text{obs} = \epsilon_b (1 + \xi), \qquad
\epsilon_b \sim \text{Unif}(0, \nu), \quad
\xi \sim \text{Unif}(-\varepsilon, \varepsilon),
$$

a product of two independent uniform random variables, not itself
uniform. Following the same conditioning argument Teunis and van
Eijkeren use to derive their combined-noise density [@Teunis_2020],
its density (the never-infected contribution to the overall observed
density, including the never-infected weight $\expf{-\lambda a}$) is

$$
\rho_{bm}(y \mid \text{never-infected}) =
\begin{cases}
\dfrac{\expf{-\lambda a}}{2\varepsilon\nu}
  \log\!\left(\dfrac{1+\varepsilon}{1-\varepsilon}\right)
  & 0 < y \le \nu(1-\varepsilon) \\[8pt]
\dfrac{\expf{-\lambda a}}{2\varepsilon\nu}
  \log\!\left(\dfrac{\nu(1+\varepsilon)}{y}\right)
  & \nu(1-\varepsilon) < y < \nu(1+\varepsilon) \\[8pt]
0 & y \ge \nu(1+\varepsilon)
\end{cases}
$$

The first piece matches Teunis and van Eijkeren's Equation 19: it is
flat (constant in $y$) over most of its support, then tapers to zero
by $y = \nu(1+\varepsilon)$.
Integrating both pieces over their full range recovers exactly
$\expf{-\lambda a}$, the never-infected probability, confirming the
density is correctly normalized.
As $\varepsilon \to 0$ the flat piece reduces to
$\expf{-\lambda a}/\nu$, the biological-noise-only density on
$[0, \nu]$, matching the single-source case above.
