# Modeling the seroresponse kinetics curve

::: notes
Now, we need a model for the antibody response to infection, $\pdf(Y=y|T=t)$.
The current version of the `{serocalculator}` package uses
a two-phase model for
the shape of the seroresponse [@Teunis_2016].
:::

## Model for active infection period {.smaller}

::: notes
The first phase of the model represents the **active infection period**,
and uses a simplified Lotka-Volterra predator-prey model
[@volterra1928variations]
where the pathogen is the prey and the antibodies are the predator:
:::

Notation:

* $x(t)$: pathogen concentration at time $t$
* $y(t)$: antibody concentration at time $t$
* $\mu_0$: pathogen growth rate; $\beta$: pathogen-inactivation strength; $\mu$: antibody growth rate

Model:

* $$x'(t) = \mu_0\, x(t) - \beta y(t)$$
* $$y'(t) = \mu\, y(t)$$

::: notes
With baseline antibody concentration $y(0) = y_{0}$ and
initial pathogen concentration
$x(0) = x_{0}$.
:::

::: notes
Compared to the standard LV model:

* the predation term with the $\beta$ coefficient is missing
the prey concentration $x(t)$ factor;
we assume that the efficiency of predation
doesn't depend on pathogen concentration.

* the differential equation for predator density is missing
the predator death rate term
$-\gamma y(t)$; we assume that as long as there are any pathogens present,
the antibody decay rate is negligible compared to the growth rate.

* the predator growth rate term $\mu\, y(t)$ is missing
the prey density factor $x(t)$;
we assume that as long as there are any pathogens present,
the antibody concentration grows at the same exponential rate.

These omissions were made to simplify the estimation process,
under the assumption that they are negligible
compared to the other terms in the model.

:::

## Model for post-infection antibody decay

::: notes
Once the immune response clears the infection, the pathogen concentration
reaches zero, so antibody production is no longer stimulated and the
antibody concentration decays.
:::

Once the pathogen is cleared ($x(t) = 0$), the antibody concentration decays:

* $$y^{\prime}(t) = -\alpha y(t)^r$$

::: notes
Antibody decay is different from exponential (log--linear) decay. 
When the shape parameter $r > 1$,
log concentrations decrease rapidly after infection has terminated,
but decay then slows down and
low antibody concentrations are maintained for a long period.
If $r = 1$, 
this model reduces to exponential decay with decay rate $\alpha$.
:::

## Assembling the full response curve

The serum antibody response $y(t)$ can be written as

$$
y(t) = y_{+}(t) + y_{-}(t)
$$

where

$$
\begin{align}
y_{+}(t) & = y_{0}\text{e}^{\mu t}[0\leq t <t_{1}]\\
y_{-}(t) & = y_{1}\left(1+(r-1)y_{1}^{r-1}\alpha(t-t_{1})\right)^{-\frac{1}{r-1}}[t_{1}\le t < \infty]
\end{align}
$$

## Growth rate from peak and baseline {visibility="hidden"}

Since the peak level is $y_{1} = y_{0}\text{e}^{\mu t_{1}}$
the growth rate $\mu$ can be written as
$$\mu = \frac{1}{t_{1}}\log\left(\frac{y_{1}}{y_{0}}\right)$$

## From the full model to the observed curve {.smaller}

::: notes
The antibody curve $y(t)$ that we observe and plot falls out of the
two-phase system as follows:
:::

* **During infection** ($t < t_1$): the antibody equation
$y'(t) = \mu\, y(t)$ gives exponential growth, $y(t) = y_0\, e^{\mu t}$.

* The **pathogen equation sets the peak time** $t_1$: circulating
antibodies inactivate the pathogen (the $-\beta y(t)$ term), driving $x(t)$
to zero at $t_1$. Once the pathogen is cleared, stimulation stops and decay
begins, so the peak antibody level is $y_1 = y_0\, e^{\mu t_1}$.

* **After the peak** ($t \ge t_1$): power-law decay,
$y'(t) = -\alpha y(t)^r$.

::: notes
So the pathogen trajectory $x(t)$ never appears in the final antibody curve
--- it enters only through the peak time $t_1$. From antibody data alone
the pathogen-side quantities ($\mu_0$, $\beta$, and the inoculum $x_0$) are
not separately identifiable [@Teunis_2016], so the curve is summarized by
five parameters: $y_0,\ y_1,\ t_1,\ \alpha,\ r$.
:::

---

```{r}
cur_ai <- "HlyE_IgG"
```

```{r}
#| label: fig-response-graph
#| fig-cap: An example kinetics curve for HlyE IgG

library(serocalculator)
library(dplyr)
# Import longitudinal antibody parameters from OSF
curves <-
  "https://osf.io/download/rtw5k/" |>
  load_sr_params() |>
  filter(iter < 50)

curve1 <-
  curves |>
  filter(
    iter == 5,
    antigen_iso == cur_ai
  )
library(ggplot2)

curve1 |>
  # plot_curve_params_one_ab() picks rows via `iter == seq_len(nrow(.))`, so a
  # single pre-filtered curve must be relabelled iter 1 or no curve is drawn.
  mutate(iter = 1) |>
  serocalculator:::plot_curve_params_one_ab(
    log_y = FALSE
  ) +
  xlim(0, 100) +
  theme_minimal() +
  geom_vline(
    aes(
      xintercept = curve1$t1,
      col = "t1"
    )
  ) +

  geom_hline(
    aes(
      yintercept = curve1$y0,
      col = "y0"
    )
  ) +


  geom_hline(
    aes(
      yintercept = curve1$y1,
      col = "y1"
    )
  ) +
  geom_point(
    data = curve1,
    aes(
      x = t1,
      y = y1,
      col = "(t1,y1)"
    )
  ) +
  theme(legend.position = "bottom") +
  labs(col = "")
```


:::: notes
The antibody level at $t=0$ is $y_{0}$;
the rising branch ends at $t = t_{1}$
where the peak antibody level $y_{1}$ is reached.
Any antibody level $y(t) \in (y_{0}, y_{1})$ eventually occurs twice.
::::

---

::: notes
An interactive Shiny app lets you manipulate the model parameters:
:::

Interactive Shiny app:
<https://ucdserg.shinyapps.io/antibody-kinetics-model-2/>

```{r}
#| echo: false
#| eval: false
#| label: qr-code
library(qrcode)
code <- qr_code("https://ucdserg.shinyapps.io/antibody-kinetics-model-2/")
generate_svg(code, filename = "man/figures/qr_akm2.svg")
```

![QR code for the antibody-kinetics Shiny app](../man/figures/qr_akm2.svg){width="280"}

## Biological noise

::: notes
When we measure antibody concentrations in a blood sample, we are essentially counting molecules (using biochemistry).

We might miss some of the antibodies (undercount, false negatives) and we also might incorrectly count some other molecules that aren't actually the ones we are looking for (overcount, false positives, cross-reactivity).

We are more concerned about overcount (cross-reactivity) than undercount. For a given antibody, we can do some analytical work beforehand to estimate the distribution of overcounts, and add that to our model $p(Y=y|T=t)$.
:::

Notation:

* $y_\text{obs}$: measured serum antibody concentration
* $y_\text{true}$: "true" serum antibody concentration
* $\epsilon_b$: noise due to probe cross-reactivity

Model:

* $y_\text{obs} = y_\text{true} + \epsilon_b$
* $\epsilon_b \sim \text{Unif}(0, \nu)$

$\nu$ needs to be pre-estimated using negative controls,
typically using the 95th percentile of the distribution of 
antibody responses to the antigen-isotype 
in a population with no exposure.

## Measurement noise

There are also some other sources of noise in our bioassays; 
user differences in pipetting technique, 
random ELISA plate effects, etc. 
This noise can cause both overcount and undercount. 
We can also estimate the magnitude of this noise source
and include it in $p(Y=y|T=t)$.

Measurement noise, $\varepsilon$ ("epsilon"), 
represents measurement error from the laboratory testing process. 
It is defined by a CV (coefficient of variation) as the ratio of the standard deviation to the mean for replicates. 
Note that the CV should ideally be measured across plates rather than within the same plate. 
