---
title: "Enteric Fever Seroincidence Vignette"
author: "UC Davis Seroepidemiology Research Group (SERG)"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    toc: true
vignette: >
  %\VignetteIndexEntry{Enteric Fever Seroincidence Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../references.bib
---
## Introduction

This vignette provides users with an example analysis using the [**serocalculator**](https://github.com/UCD-SERG/serocalculator) package 
by reproducing the analysis for: 
[**Estimating typhoid incidence from community-based serosurveys: a multicohort study**](https://www.thelancet.com/journals/lanmic/article/PIIS2666-5247(22)00114-8/fulltext) (@Aiemjoy_2022_Lancet). 
We review the methods underlying the analysis 
and then walk through an example of enteric fever incidence in Pakistan. 
Note that because this is a simplified version of the analysis, the results here will differ slightly from those presented in the publication.

In this example, users will determine the seroincidence of enteric fever in cross-sectional serosurveys conducted as part of the SeroEpidemiology and Environmental Surveillance (SEES) for enteric fever study in Bangladesh, Nepal, and Pakistan. Longitudinal antibody responses were modeled from 1420 blood culture-confirmed enteric fever cases enrolled from the same countries. 


```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height = 4,
  fig.width = 7,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```
## Methods

### Load packages

The first step in conducting this analysis is to load our necessary packages. 
**If you haven't installed already, you will need to do so before loading:**

```{r, eval = FALSE}
install.packages("serocalculator")
```

See the 
[Installation instructions](https://ucd-serg.github.io/serocalculator/#installing-the-serocalculator-package) for more details.

Once the `serocalculator` package has been installed, 
load it into your R environment using `library()`,
along with any other packages you may need for data management;
For this example, we will load `tidyverse` and `forcats`:

```{r load_packages}
library(serocalculator)
library(tidyverse)
library(forcats)
``` 

## Load data

Pathogen-specific sample datasets, noise parameters, and longitudinal antibody dynamics for **serocalculator** are available on the [Serocalculator Data Repository](https://osf.io/ne8pc/) on Open Science Framework (OSF). We will pull this data directly into our R environment.

Note that each dataset has specific formatting and variable name requirements.


### Load and prepare longitudinal parameter data

We will first load the longitudinal curve parameters to set the antibody decay parameters. In this example, these parameters were modeled with Bayesian hierarchical models to fit two-phase power-function decay models to the longitudinal antibody responses among confirmed enteric fever cases. 

*Formatting Specifications*: Data should be imported as a "wide" dataframe with one column for each parameter and one row for each iteration of the posterior distribution for each antigen isotype. Column names must *exactly* match follow the naming conventions:

Column Name    |  Description
-------------- |  -----------
y0         |  Baseline antibody concentration
y1         |  Peak antibody concentration
t1         |  Time to peak antibody concentration
alpha      |  Antibody decay rate
r          |  Antibody decay shape
*Note that variable names are case-sensitive*


```{r curve, message=FALSE}
# Import longitudinal antibody parameters from OSF
curves <-
  "https://osf.io/download/rtw5k/" |>
  load_sr_params()
```

### Visualize curve parameters

We can graph the decay curves with an `autoplot()` method:

```{r, fig.width=7, fig.height=4}
# Visualize curve parameters
curves |>
  filter(
    antigen_iso %in% c("HlyE_IgA", "HlyE_IgG")
  ) |>
  autoplot()

```


### Load and prepare cross-sectional data 
Next, we load our sample cross-sectional data. We will use a subset of results from the SEES dataset. Ideally, this will be a representative sample of the general population without regard to disease status. Later, we will limit our analysis to cross-sectional data from Pakistan. 

We have selected hemolysin E (*HlyE*) as our target antigen and *IgG* and *IgA* as our target immunoglobulin isotypes. Users may select different serologic markers depending on what is available in your data. From the original dataset, we rename our result and age variables to the names required by **serocalculator**. 

*Formatting Specifications*: Cross-sectional population data should be a "long" dataframe with one column for each variable and one row for each antigen isotype resulted for an individual. So the same individual will have more than one row if they have results for more than one antigen isotype. The dataframe can have additional variables, but the two below are required: 


Column Name | Description
----------- | -----------
value     | Quantitative antibody response
age       | Numeric age
*Note that variable names are case sensitive*


```{r data, message = FALSE}
#Import cross-sectional data from OSF and rename required variables
xs_data <- readr::read_rds("https://osf.io/download//n6cp3/") |>
  as_pop_data()

```

### Summarize antibody data

We can compute numerical summaries of our cross-sectional antibody data with a `summary()` method for `pop_data` objects:

```{r}
xs_data |> summary()


```


### Visualize antibody data

We examine our cross-sectional antibody data by visualizing the distribution of quantitative antibody responses. Here, we will look at the distribution of our selected antigen and isotype pairs, HlyE IgA and HlyE IgG, across participating countries.

```{r plots}

#color palette
country_pal <- c("#EA6552", "#8F4B86", "#0099B4FF")

xs_data |> autoplot(strata = "Country", type = "density") +
  scale_fill_manual(values = country_pal)

```

We see that across countries, our data is highly skewed with the majority of responses on the lower end of our data with long tails. Let's get a better look at the distribution by log transforming our antibody response value. 

```{r logplot, message = FALSE}
# Create log transformed plots

xs_data |>
  mutate(Country = fct_relevel(Country, "Bangladesh", "Pakistan", "Nepal")) |>
  autoplot(strata = "Country", type = "density") +
  scale_fill_manual(values = country_pal) +
  scale_x_log10(labels = scales::label_comma())

```

Once log transformed, our data looks much more normally distributed. In most cases, log transformation will be the best way to visualize serologic data. 

Let's also take a look at how antibody responses change by age.

```{r plot-age, message=FALSE, warning = FALSE}

#Plot antibody responses by age
xs_data |>
  autoplot(
    strata = "Country",
    type = "age-scatter"
  ) +
  scale_color_manual(values = country_pal)

```

In this plot, we can see that the highest burden is in Bangladesh, but Nepal has the steepest slope and experiences the greatest change in seroconversion rates by age, with lower exposures at younger ages and higher exposures at older ages.

### Load noise parameters

Next, we must set conditions based on some assumptions about the data and errors that may need to be accounted for. This will differ based on background knowledge of the data. 


The biological noise, $\nu$ ("nu"), represents error from cross-reactivity to other antibodies. It is defined as the 95th percentile of the distribution of antibody responses to the antigen-isotype in a population with no exposure. 

Measurement noise, $\varepsilon$ ("epsilon"), represents measurement error from the laboratory testing process. It is defined by a CV (coefficient of variation) as the ratio of the standard deviation to the mean for replicates. Note that the CV should ideally be measured across plates rather than within the same plate. 

*Formatting Specifications*: Noise parameter data should be a dataframe with one row for each antigen isotype and columns for each noise parameter below.


Column Name | Description
----------- | -----------
y.low       | Lower limit of detection of the assay
nu          | Biologic noise
y.high      | Upper limit of detection of the assay
eps         | Measurement noise
*Note that variable names are case-sensitive.*

``` {r noise, message=FALSE}
# Import noise parameters from OSF

noise <- url("https://osf.io/download//hqy4v/") |> readRDS()
```

## Estimate Seroincidence 
Now we are ready to begin estimating seroincidence. We will conduct two separate analyses using two distinct functions, `est.incidence` and `est.incidence.by`, to calculate the overall seroincidence and the stratified seroincidence, respectively.

### Overall Seroincidence
Using the function `est.incidence`, we filter to sites in Pakistan and define the datasets for our cross-sectional data (pop_data), longitudinal parameters (curve_param), and noise parameters (noise_param). We also define the antigen-isotype pairs to be included in the estimate (antigen_isos). Here, we have chosen to use two antigen isotypes, but users can add additional pairs if available. 

```{r est}
# Using est.incidence (no strata)

est1 <- est_seroincidence(
  pop_data = xs_data |> filter(Country == "Pakistan"),
  sr_params = curves,
  noise_params = noise |> filter(Country == "Pakistan"),
  antigen_isos = c("HlyE_IgG", "HlyE_IgA")
)

summary(est1)
```


### Stratified Seroincidence

We can also produce stratified seroincidence estimates. Users can select one or more stratification variables in their cross-sectional population dataset. 
Let's compare estimates across all countries and by age group. 

```{r estbycountry}
#Using est.incidence.by (strata)

est_country_age <- est_seroincidence_by(
  strata = c("Country", "ageCat"),
  pop_data = xs_data,
  sr_params = curves,
  curve_strata_varnames = NULL,
  noise_params = noise,
  noise_strata_varnames = "Country",
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
  num_cores = 8 # Allow for parallel processing to decrease run time
)

summary(est_country_age)

```
Note that we get a warning about uneven observations between antigen isotypes, meaning some participants do not have results for both *HlyE IgA* and *HlyE IgG*. 
The warning indicates that the "*Sample size for each stratum will be calculated as the minimum number of observations across all antigen isotypes*", so only participants with both antigen isotypes are included. 
To avoid this, filter the dataset to include only records with all specified antigen isotypes.

We set `curve_strata_varnames = NULL` to avoid stratification in the "curves" dataset because it does not include our strata variables. 
Without this, a warning appears: "*curve_params is missing all strata variables, and will be used unstratified*". 
To stratify based on variables that exist in a longitudinal curve parameters dataset, specify variables using `curve_strata_varnames`, similar to how `noise_strata_varnames` is used for "noise" above.


Finally, let's visualize our seroincidence estimates by country and age category using `autoplot()`:

```{r}
# Save summary(est_country_age)
est_country_agedf <- summary(est_country_age)

# Plot seroincidence estimates
autoplot(
  est_country_agedf,
  type = "bar",
  yvar = "ageCat",
  color_var = "Country",
  color_palette = country_pal,
  CIs = TRUE
)
```


## Comparing Seroincidence Rates

After estimating seroincidence rates across different groups, we may want to statistically compare these rates to determine if observed differences are significant. The `compare_seroincidence()` function performs two-sample z-tests to compare seroincidence rates.

### Comparing Two Individual Estimates

First, let's compare seroincidence rates between two countries directly. We'll estimate seroincidence for Bangladesh and Nepal separately, then compare them:

```{r compare_two}
# Estimate seroincidence for Bangladesh
est_bangladesh <- est_seroincidence(
  pop_data = xs_data |> filter(Country == "Bangladesh"),
  sr_params = curves,
  noise_params = noise |> filter(Country == "Bangladesh"),
  antigen_isos = c("HlyE_IgG", "HlyE_IgA")
)

# Estimate seroincidence for Nepal
est_nepal <- est_seroincidence(
  pop_data = xs_data |> filter(Country == "Nepal"),
  sr_params = curves,
  noise_params = noise |> filter(Country == "Nepal"),
  antigen_isos = c("HlyE_IgG", "HlyE_IgA")
)

# Compare the two estimates
comparison <- compare_seroincidence(est_bangladesh, est_nepal)
print(comparison)
```

The output follows the standard `htest` format in R, providing the z-statistic, p-value, estimates for both groups, and a confidence interval for the difference in incidence rates.

### Comparing All Pairs of Stratified Estimates

For stratified analyses, we can compare all pairs of strata at once. This is particularly useful when we have multiple groups and want a comprehensive view of all pairwise differences:

```{r compare_strata}
# Compare all pairs of country-age combinations
comparisons_table <- compare_seroincidence(est_country_age)

# Display the results
print(comparisons_table)
```

This produces a table with one row for each pairwise comparison, showing:
- The two strata being compared
- Incidence rates for each group
- The difference in rates
- Standard error of the difference
- Z-statistic
- P-value
- 95% confidence interval for the difference

We can use this to identify which differences are statistically significant (typically using p < 0.05 as a threshold).

## Accounting for Clustered Sampling Designs

In many survey designs, observations are clustered (e.g., multiple individuals from the same household, geographic area, or school). When observations within clusters are more similar to each other than to observations from different clusters, standard errors that ignore clustering will be too small, leading to overconfident inference.

The `serocalculator` package supports cluster-robust standard error estimation through the `cluster_var` parameter.

### Example: Household Clustering

The SEES data includes cluster identifiers (`cluster` for geographic clusters, `catchment` for health facility catchment areas, and `Country`). These variables are nested: clusters are nested within catchments, which are nested within countries. We can use these to account for within-cluster correlation:

```{r clustering_example}
# Account for geographic clustering
# Clusters in the SEES data represent geographic areas
est_with_clustering <- est_seroincidence(
  pop_data = xs_data,
  sr_params = curves,
  noise_params = noise,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
  cluster_var = "cluster"  # Geographic cluster identifier
)

# The summary will show cluster-robust standard errors
summary(est_with_clustering)
```

### Multi-level Clustering

For the SEES data with nested clustering structure (clusters within catchments, catchments within countries), we can specify multiple cluster variables:

```{r multilevel_clustering}
# Multi-level clustering: clusters nested within catchments
# Note: This assumes 50% of clusters in each catchment were sampled
est_multilevel <- est_seroincidence(
  pop_data = xs_data,
  sr_params = curves,
  noise_params = noise,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
  cluster_var = c("catchment", "cluster")  # Nested clustering
)

summary(est_multilevel)
```

### Clustering with Stratified Analysis

Clustering can also be combined with stratified analysis using `est_seroincidence_by()`:

```{r clustering_stratified}
# Stratified analysis by catchment with cluster adjustment
est_catchment_clustered <- est_seroincidence_by(
  pop_data = xs_data,
  strata = "catchment",
  sr_params = curves,
  noise_params = noise,
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"),
  cluster_var = "cluster"  # Account for clustering within each stratum
)

summary(est_catchment_clustered)
```

### Understanding the Impact

When cluster-robust standard errors are used:

- **Point estimates** (incidence rates) remain unchanged
- **Standard errors** typically increase by 5-15% to reflect within-cluster correlation
- The `summary()` output includes a `se_type` column indicating "cluster-robust" vs "standard"
- Confidence intervals appropriately widen to account for reduced effective sample size


```{r include=FALSE}
#Calculate output values
rate_bangla_5_15 <- est_country_agedf |> 
  dplyr::filter(Country == "Bangladesh", ageCat == "5-15") |> 
  pull("incidence.rate") 

rate_bangla_5_15 <- round(rate_bangla_5_15 * 1000)

rate_nepal_5_15 <- est_country_agedf |> 
  dplyr::filter(Country == "Nepal", ageCat == "5-15") |> 
  pull("incidence.rate") 

rate_nepal_5_15 <- round(rate_nepal_5_15 * 1000)

rate_ratio_bangla_nepal <- round(rate_bangla_5_15 / rate_nepal_5_15)
```

## Conclusions
We estimate that Bangladesh has the highest enteric fever seroconversion rates across all age groups, with the highest rates observed among 5- to 15-year-olds (`r rate_bangla_5_15` per 1000 person-years). In this age group, the seroconversion rate in Bangladesh is `r rate_ratio_bangla_nepal` times higher than in Nepal, where the rate is `r rate_nepal_5_15` per 1000 person-years. These findings highlight substantial geographic variation in enteric fever transmission, emphasizing the need for targeted prevention strategies. **serocalculator** offers an efficient and reproducible approach to estimating seroconversion rates, enabling data-driven insights for disease surveillance and public health decision-making.


## Acknowledgments
We gratefully acknowledge the study participants for their valuable time and interest in participating in these studies. Special thanks to our collaborators at Sabin Vaccine Institute, Aga Khan University  (Karachi, Pakistan), Child Health Research Foundation (Dhaka, Bangladesh), and Dhulikhel Hospital, Kathmandu University Hospital (Dhulikhel, Nepal). 

## Funding
This project was supported by grants from the National Institutes of Health (NIH) National Institute of Allergy and Infectious Disease (R21AI176416), the NIH Fogarty International Center (K01TW012177) and the Bill and Melinda Gates Foundation. 

## References

