---
title: "Pertussis Seroconversion Rate Vignette"
author: "UC Davis Seroepidemiology Research Group (SERG) and Peter Teunis"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    toc: true
vignette: >
  %\VignetteIndexEntry{Pertussis Seroconversion Rate Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../references.bib
---

```{r, echo = FALSE}
pkgUrl <- gsub("\n", "", packageDescription("serocalculator")$URL)
```

# 1. Introduction

Antibody levels measured in a (cross–sectional) population sample can be translated into an estimate of the frequency with which seroconversions (infections) occur in the sampled population. In simple 
terms: the presence of many high antibody concentrations indicates that many subjects likely 
experienced infection recently, while mostly low concentrations indicate a low frequency of 
infections in the sampled population. 

In order to thus interpret the measured cross-sectional antibody concentrations, characteristics of 
the time course of the serum antibody response must be known. For any antibody detected in the 
cross–sectional sample, a quantitative description of its dynamics following infection 
(seroconversion to an elevated concentration, followed by a gradual decrease) must be available. In 
published studies, this information on the time course of the serum antibody response has been 
obtained from follow–up data in subjects who had a symptomatic episode following infection. The 
onset of symptoms thus provided a proxy for the time that infection occurred. Care must be taken 
that longitudinal and cross–sectional measurements of antibody concentrations are calibrated on the 
same scale (or, preferably, expressed in identical units).

The time course of the serum antibody response to infection is assumed generic: anyone infected by 
the same pathogen is expected to produce a response similar to those seen in the symptomatic cohort 
used for the follow–up study. Thus, the strategy is to set up a longitudinal model from a follow–up 
study once, and use the thus obtained information to analyze many different cross–sectional data 
sets.

Note that it must be assumed that such variation in the time course of antibody concentrations among individuals as observed in the longitudinal cohort is the same as in the population where the 
cross-sectional sample was taken, where this variation cannot be observed.

Analysis of longitudinal data and estimation of the required parameters is separated from the 
estimation of seroincidences, because (1) there are few longitudinal data sets available and they 
are difficult (and expensive) to obtain, (2) such longitudinal models may be considered *generic*, 
presumably describing a physiological phenomenon that is similiar in any infected individual, and 
(3) setting up and running the longitudinal model requires additional skills, not necessary for use 
of the sero–incidence calculator script

# 2. The seroconversion rate estimator

Package **serocalculator** was designed to calculate incidence of seroconversion, by using the 
longitudinal seroresponse characteristics. The distribution of serum antibody concentrations in a 
cross–sectional population sample is calculated, as a function of the (longitudinal) seroresponse 
and the frequency of seroconversion (or seroincidence). Given the seroresponse, this (marginal) 
distribution of antibody concentrations can be fitted to the cross-sectional data, by adjusting the 
seroconversion frequency, thus providing a means to estimate the seroconversion frequency.

## 2.1 Time course of serum antibodies

The seroresponse, the time course of serum antibodies following seroconversion (infection) is 
described by means of a set of parameters describing it. These parameters may include:

a) baseline antibody concentration, 
b) time to reach peak antibody concentration, 
c) magnitude of the peak concentration, and 
d) antibody decay parameters describing the time course (shape and rate) of antibody decay. 

A model description that is used for the serocalculator is described in @Simonsen_2009 and @Teunis_2012. An alternative 
model, simplified but with improved interpretation of the underlying biological mechanisms, is given 
in @deGraaf_2014 and further improved in @Teunis_2016 and @Teunis_2020.

The current model assumes that upon exposure to an initial inoculum, pathogen concentrations 
increase exponentially. Presence of pathogens (antigen presented to the immune system) signals an 
increase in antibody production, resulting in exponentially increasing serum antibody 
concentrations. Increased activity of the immune system, as signalled by circulating antibodies, 
causes inactivation or die–off of pathogens, with a rate proportional to the serum antibody 
concentration. Ultimately, when all pathogens are removed, infection is over and the promotion of 
antibody production stops. Then the decay phase starts, and serum antibody concentrations decrease. 
Antibody decay in the serum compartment may start fast and slow down later, resulting in 
non–exponential decay. In @Teunis_2016 this is modeled using a shape parameter, allowing gradual adjustment 
between exponential (log–linear) and strongly non–exponential decay.

## 2.2 Heterogenity

The five parameters describing the seroresponse (in round brackets names of corresponding 
parameters used in the package, see [section 3.2.2](#longParams) for details): baseline antibody 
concentration (`y0` and `yb`), time to peak (`t1`), peak antibody concentration (`y1`), decay rate 
(`alpha`) and shape factor of the decay curve (`r`) are not fixed numbers. In a population of 
subjects, individuals each have their own seroresponse, different from those of other subjects. 
Therefore, these parameters are provided as a (joint) distribution, describing their variation among individuals in the sampled population. Of course, such a distribution depends on the pathogen, the 
detected antibody, and even the detection assay.

Estimates for these seroresponse parameters may be obtained by fitting the model to longitudinal 
data. Typically, such data are collected in patients with confirmed infection, where the time of 
infection is detected by appearance of symptoms. Ideally, blood samples are taken at several times 
post infection, up to a period long enough to reliably estimate the shape of the serum antibody 
decay curve.

A Bayesian hierarchical framework allows for individual variation in seroresponses, so that one 
obtains the (posterior) joint distribution of the parameters as a (Markov chain) Monte Carlo sample, of all relevant parameters @Teunis_2016.

Note that, due to the limited number of iterations in the Monte Carlo sample, the generated 
incidence estimates will show minor variations: if the same calculation is run twice, the outcomes 
may differ slightly. If this variation should be smaller, the size of the Monte Carlo sample may be 
increased. This however will slow down the calculations.

## 2.3 Seroconversion rate estimation

The longitudinal seroresponses are considered generic: they represent the time course of (specific) 
human serum antibodies against a (specific) pathogen. Therefore, with any cross–sectional sample of 
the same antibodies in a human population, an incidence estimate can be calculated. 

Given the longitudinal response characteristics, **serocalculator** R package provides a rapid and 
computationally simple method for calculating seroconversion rates, as published in @Teunis_2012.

The methods on which this package is based were developed with publicly funded 
research. All procedures have been published and are open to the general public, including the core 
script on which the **sercalculator** is based. To make the functionalities provided as 
broadly available as possible, R was chosen as the computing platform, because it is free and open 
source, and runs on any modern computer operating system.

## 2.4 Model variants

The package contains longitudinal models for seroincidence calculations. Four model variants are 
possible, depending on whether seroconversion is instantaneous ($t_{1} = 0$) or not ($t_{1} > 0$), 
and whether decay is exponential ($r = 1$) or proceeds as a power function ($r > 1$). Power function 
decay allows for rapid inital decay followed by a sustained period of very slow decay (@Teunis_2016). 
Available model variants are described in Table 1.

\begin{table}[!h]
\centering
\caption{Model variants}
\vspace{0.3cm}
\begin{footnotesize}
\begin{tabular}{ccc}
 & Time to & Decay\\
Model & Seroconversion ($t_{1}$) & parameter ($r$)\\
1 & 0 & 1\\
2 & 0 & $> 1$\\
3 & $> 0$ & 1\\
4 & $> 0$ & $> 1$\\
\end{tabular}
\begin{tabular}{lllcccc}
 & & & \multicolumn{4}{c}{Model}\\
Pathogen & Antibody & Units & 1 & 2 & 3 & 4\\ \hline
{\em Campylobacter jejuni} & IgG & ELISA (Delft) & * & & * & * \\
 & IgM & ELISA (Delft) & * & & * & * \\
 & IgA & ELISA (Delft) & * & & * & * \\
 & IgG & ELISA (SSI)   & * & * & & * \\
 & IgM & ELISA (SSI)   & * & * & & * \\
 & IgA & ELISA (SSI)   & * & * & & * \\ \hline
{\em Salmonella} & IgG & mixed ELISA (SSI) & * & * & & * \\
 & IgM & mixed ELISA (SSI) & * & * & & * \\
 & IgA & mixed ELISA (SSI) & * & * & & * \\ \hline
{\em Bordetella pertussis} & IgG-PT & IU    & * & * & * & * \\ \hline
\end{tabular}
\end{footnotesize}
\end{table}

# 3. How to use *serocalculator* package

This section provides step-by-step directions on the usage of the *serocalculator* package.

## 3.1. Loading the package

Functionality of the *serocalculator* package is made available to the end user only once the library is loaded into the current workspace. Assuming the package is installed already (covered in 
[installation.pdf](installation.pdf)) loading is achieved by running the following command in the R 
console (bear in mind that the text after character `#` is only a comment):

```{r}
# Load package "serocalculator"
library(serocalculator)
```

At this moment all functions required to run the seroconversion rate calculation are made available. It can be checked by running:

```{r, eval=FALSE}
# List all objects (functions and data) exposed by package "serocalculator".
ls("package:serocalculator")
```

Here is the resulting output:

```{r, echo=FALSE}
ls("package:serocalculator")
```

We briefly describe the most important objects and functions:

a) `[DISEASE]Params[MODEL_TYPE]`: Monte Carlo sample of longitudinal response parameters per 
antibody for various diseases: Campylobacter, Pertussis, Salmonella. Object class: `list`.
b) `[DISEASE]Sim[Low|Medium|High]Data`: Example simulated antibody levels data measured in a 
cross-sectional population for three values of lambda (incidence): 0.036 ("Low"), 0.21 ("Medium") 
and 1.15 ("High") (1/yr). Object class: `data.frame`.
c) `getAdditionalData`: Utility function for downloading additional longitudinal response 
parameters from an online repository. Files available for download are `coxiellaIFAParams4.zip` and 
`yersiniaSSIParams4.zip`. Object class: `function`.
d) `estimateSeroincidence`: Main function of the package. Estimates seroincidence based on supplied 
cross-section antibody levels data and longitudinal response parameters. Object class: `function`.

## 3.2. Specifying input data

There are two sets of inputs to the serology calculator that must always be specified:

a) Simulated antibody levels measured in cross-sectional population sample (see 
`campylobacterSimLowData`, `campylobacterSimMediumData`, etc. above).
b) Longitudinal response characteristic as set of parameters `(y1, alpha, yb, r, y0, mu1, t1)`
(see `campylobacterDelftParams1`, `campylobacterSSIParams2`, etc. above).

We start with loading antibody levels measured in cross-sectional population sample.

### 3.2.1. Specifying cross-sectional antibody levels data

The user has a few options of providing this data to the calculator:

**i) Using supplied sample data**

Data set `campylobacterSimMediumData` provides an example of a cross-sectional antibody levels data 
for Campylobacter. It is loaded into the workspace as soon as the package is loaded, therefore it
can be referenced right away. Here is a small exempt from this data set:

```{r}
# Show three first rows of "campylobacterSimMediumData" data.frame.
head(campylobacterSimMediumData, 3)
```

Let us create the serology data `serologyData` based on this data set:
```{r}
# Assign data.frame "campylobacterDelftData" to object named "serologyData".
serologyData <- with(campylobacterSimMediumData,
                     data.frame(
                       Age = Age,
                       IgG = IgG,
                       IgM = IgM,
                       IgA = IgA,
                       AgeCat = cut(Age, 15, labels = FALSE)))

# Print a few first observations.
head(serologyData)
```

You can view the data also in a tabular form:
```{r, eval=FALSE}
View(serologyData)
```

We shall pass this object later to the calculator. Object `serologyData` must contain at least one 
column named `IgG`, `IgM`, or `IgA`. Additionally, column `Age` is used in the calculations if it is 
present. Otherwise, it is created on-the-fly and initialized to `NA` (not available). Column `Age`,
if supplied, is assumed to represent age in days.

Please, note that the cross-sectional data set can include extra columns allowing performing the 
seroincidence calculations stratified per factors available in those data. In the example above 
column `AgeCat` can be used as a *stratum* variable. This particular variable is created by
mapping the age in days (column `Age`) into 15 categories, category 1 grouping the youngest subjects 
and category 15 grouping the oldest subjects.

**ii) Loading from external files**

The user can load her/his own data into an R session. Details on loading specific file types can be 
found in online R manual on [R Data Import/Export](http://cran.r-project.org/doc/manuals/r-release/R-data.html).

It is important that the input data is in a tabular form with each column containing levels measured 
for a single antibody type. Column name should indicate name of the measured antibody. For instance, 
a comma separated file with the following content:

```
IgG, IgM, IgA
5.337300, 0.4414653, 0.3002395
3.534118, 0.3888226, 0.3543486
2.144549, 0.3320178, 0.3884205
2.957854, 0.4967764, 0.3472340
```

is a valid cross-sectional data set. Suppose this file is named `c:\cross-sectional-data.csv`.
Then it can be loaded into R like this:
```{r, eval=FALSE}
# Read content of file "C:\cross-sectional-data.csv" into object named 
# "serologyData".
serologyData <- read.csv(file = "C:\\cross-sectional-data.csv")
```

### 3.2.2. Specifying longitudinal response parameters {#longParams}

The longitudinal response parameters set consists of the following items:

- `y1`: antibody peak level (ELISA units)
- `alpha`: antibody decay rate (1/days for the current longitudinal parameter sets)
- `yb`: baseline antibody level at $t$ approaching infinity ($y(t->inf)$)
- `r`: shape factor of antibody decay
- `y0`: baseline antibody level at $t=0$ ($y(t=0)$)
- `mu1`: initial antibody growth rate
- `t1`: duration of infection

Please, refer to vignette [methodology.pdf](methodology.pdf) for more details on the underlying 
methodology.

End user has three options for loading the response parameters into R session:

**i) Using supplied data**

There are a few longitudinal response parameter data sets provided by the package including 
Monte-Carlo sample of longitudinal response parameters, like `campylobacterDelftParams4`.
This particular object is a list containing three data.frame objects named `IgG`, `IgM` and `IgA`.
Let's have a pick into each of these two data sets separately (notice character `$` indicating a 
sub-object of the parent object):
```{r}
# Show first rows of data.frame "IgG" in list "campylobacterDelftParams4".
head(campylobacterDelftParams4$IgG)

# Show first rows of data.frame "IgM" in list "campylobacterDelftParams4".
head(campylobacterDelftParams4$IgM)

# Show first rows of data.frame "IgA" in list "campylobacterDelftParams4".
head(campylobacterDelftParams4$IgA)
```

Let us assign object `campylobacterDelftParams4` to variable `responseParams`:
```{r}
responseParams <- campylobacterDelftParams4
```

**ii) Loading from local external files**

Alternatively, this data can be loaded from an external file. Bear in mind, that most likely
your data is organized into three files, separately for antibody `IgG`, `IgM` and `IgA`. Assuming
that the data is saved into csv files named `IgG.csv`, `IgM.csv` and `IgA.csv` the end user would 
run the following code in order to create a valid input to the calculator:
```{r, eval=FALSE}
IgGData <- read.csv(file = "IgG.csv")
IgMData <- read.csv(file = "IgM.csv")
IgAData <- read.csv(file = "IgA.csv")

# Create a list named "responseData" containing objects named "IgG", "IgM" and 
# "IgA".
responseParams <- list(IgG = IgGData, IgM = IgMData, IgA = IgAData)
```

Note that object `responseParams` is a list with three dataframes named `IgG`, `IgM` and `IgA` as 
required.

**iii) Loading from an external repository**

Finally, response parameters can be obtained from an online repository set up for this package. The
repository is hosted on ECDC servers. At the time of publishing this package two files are 
available:

a) `coxiellaIFAParams4.zip`: longitudinal response parameters per antibody for Coxiella.
b) `yersiniaSSIParams4.zip`: longitudinal response parameters per antibody for Yersinia.

Simply use the supplied utility function `getAdditionalData` to download this data:
```{r, eval=FALSE}
responseParams <- getAdditionalData("coxiellaIFAParams4.zip")
```

Internet connection is needed for this function to work.

## 3.3. Estimating seroconversion rate

Main calculation function provided by package *serocalculator* is called `est.incidence`.
It takes several arguments:

- `data`: A data frame with the cross-sectional data (see variable `serologyData` above). This may 
  be the whole data set loaded from file, but can also be a subset, e.g. `data = serologyData[1, ]` 
  to use only the first row. In this tutorial this is the data loaded into variable `serologyData`. 
  It is a required input.
- `antibodies`: A list of antibodies to be used. This can be a triplet 
  (`antibodies = c("IgG", "IgM", "IgA")`) or any single antibody (`antibodies = c("IgG")`) or a 
  combination of any two antibodies. It is a required input.
- `strata`: A list of categories to stratify data. The column `AgeCat` in the example data set can 
  be used as an example. If `strata = ""` (default), then the whole data set is treated as a single 
  stratum; if each sample is assigned a unique identifier (e.g. `strata = "AgeCat"`) there are as 
  many strata as there are samples and an incidence estimate is calculated for each sample. If not 
  specified, then `strata` is initialized to `""`.
- `params`: A list with the longitudinal parameters for all antibodies specified. In this tutorial 
  this is the data loaded into variable `responseParams`. It is a required input.
- `censorLimits`: A list of cutoff levels, one for each antibody used. These are antibody levels 
  below which observations should be treated as censored. It is a required input.
- `par0`: List of parameters for the (lognormal) distribution of antibody concentrations for true 
  seronegatives (i.e. those who never seroconverted), by named antibody type (corresponding to 
  `data`).
- `start` A starting value for `log(lambda)`. Value of -6 corresponds roughly to 1 day 
  (`log(1/365.25)`), value of -4 corresponds roughly to 1 week (`log(7/365.25)`). Users are adivised 
  to experiment with this value to confirm that convergence to the same estimate is obtained. 
  Default is -6.
- `numCores` Number of processor cores to use for calculations when computing by `strata`. If set to 
  more than 1 and R package `parallel` is available (installed by default), then the computations by 
  stratum are executed in parallel. Default is 1, ie. no execution in parallel.

Here is an example of the *Campylobacter* seroincidence calculation for three antibodies measured,
stratified per `AgeCat`, and with measurements below `0.1` removed:
```{r, eval=FALSE}
# Prepare input data.
serologyData <- with(campylobacterSimMediumData,
                     data.frame(
                       Age = rep(NA, nrow(campylobacterSimMediumData)),
                       IgG = IgG,
                       IgM = IgM,
                       IgA = IgA,
                       AgeCat = cut(Age, 15, labels = FALSE)))

responseParams <- campylobacterDelftParams4

cutOffs <- list(IgG = 0.1, IgM = 0.1, IgA = 0.1)

# Baseline distributions: the distributions of antibody concentrations in 
# subjects who have never seroconverted.
baseLine <- list(IgG = c(log(0.05), 1),
                 IgM = c(log(0.005), 1),
                 IgA = c(log(0.005), 1))

# Assign output of function "estimateSeroincidence" to object named 
# "seroincidenceData". Use all available processor cores.
seroincidenceData <- estimateSeroincidence(
  data = serologyData,
  antibodies = c("IgG", "IgM", "IgA"),
  strata = "AgeCat",
  params = responseParams,
  censorLimits = cutOffs,
  par0 = baseLine,
  start = -4,
  numCores = parallel::detectCores())

# Show content of the output variable...
print(seroincidenceData)
# ...or simply type in the console: 'seroincidenceData' (without "'") and 
# press ENTER.
```

The following text should be printed.

    ## Seroincidence object estimated given the following setup:
    ## a) Antibodies   : IgG, IgM, IgA
    ## b) Strata       : AgeCat 
    ## c) Censor limits:  IgG = 0.1, IgM = 0.1, IgA = 0.1 
    ## 
    ## This object is a list containing the following items:
    ## Fits         - List of outputs of "optim" function per stratum.
    ## Antibodies   - Input parameter antibodies of function "estimateSeroincidence".
    ## Strata       - Input parameter strata of function "estimateSeroincidence".
    ## CensorLimits - Input parameter censorLimits of function "estimateSeroincidence".
    ## 
    ## Call summary function to obtain output results.

The most important output is subobject `Fits` containing raw results on the output incidence. 
Translation of these raw results is provided by a custom function `summary` explained in the 
following section.

The cutoff argument is based on censoring of the observed serum antibody measurements @Strid_2001. Cut-off levels must always be specified when calling function `estimateSeroincidence` (argument 
`censorLimits`). Value `0` should be set for antibody measurements where no censoring is needed, for 
instance 

```{r, eval=FALSE}
censorLimits <- list(IgG = 0, IgM = 0, IgA = 0)
```

Using different censoring levels will produce different results. Choosing higher cut-off causes the 
estimates of lambda to decrease. Results should be published together with the chosen cut-off 
values. The cut-off level of `0.1` set in the example above is a typical value, but users should set 
it to value reflecting the censoring level in the data.

## 3.4. Getting a summary of seroconversion rate

Output of the calculator should be passed in to function `summary` calculating output results:

```{r, eval=FALSE}
summary(seroincidenceData)
```

    ## Seroincidence estimated given the following setup:
    ## a) Antibodies   : IgG, IgM, IgA 
    ## b) Strata       : AgeCat 
    ## c) Censor limits: IgG = 0.1, IgM = 0.1, IgA = 0.1 
    ## d) Quantiles    : 0.025, 0.975 
    ## 
    ##  Seroincidence estimates:
    ##    Lambda.est Lambda.lwr Lambda.upr  Deviance Convergence Stratum
    ## 1  0.09721890 0.06397004 0.14774909 105.29815           0       1
    ## 2  0.05497303 0.03573489 0.08456817 181.90627           0       2
    ## 3  0.04456096 0.02845903 0.06977327 199.99357           0       3
    ## 4  0.02944038 0.01715302 0.05052964 141.32558           0       4
    ## 5  0.02569781 0.01436524 0.04597053 123.33213           0       5
    ## 6  0.02565297 0.01587275 0.04145941 187.41060           0       6
    ## 7  0.03458616 0.02012722 0.05943207 128.84486           0       7
    ## 8  0.02776873 0.02640621 0.02920154 133.57166          52       8
    ## 9  0.04016261 0.02133209 0.07561541  94.99389           0       9
    ## 10 0.01996477 0.01898409 0.02099612 104.30992          52      10
    ## 11 0.03356017 0.01917771 0.05872886 141.20168           0      11
    ## 12 0.05287648 0.03263923 0.08566140 181.68655           0      12
    ## 13 0.05081097 0.02946963 0.08760729 123.66670          52      13
    ## 14 0.04813104 0.02945312 0.07865371 186.92218           0      14
    ## 15 0.03421042 0.02210895 0.05293569 215.78684           0      15

The `summary` function returns a list of objects as well. It contains sub-objects `Antibodies`, 
`Strata`, `CensorLimits`, `Quantiles` and `Results`.

The most important sub-object of the output of the `summary` functions is `Results`, containing 
estimates of `Lambda` (annual incidence) together with their lower and upper bounds (`Lambda.lwr` 
and `Lambda.upr`, respectively). The default lower bound is 2.5% and the upper bound is 97.5% of the 
distribution of `Lambda`. These values can be changed (see [Other examples](#other-examples)).

Optionally, the summary function returns also `Deviance` (negative log likelihood at the estimated 
`Lambda`) and `Convergence` (indicator returned by function `stats::optim`; value of 0 indicates 
convergence).

It may be beneficial to assign output of this function to a variable for further analysis:

```{r, eval=FALSE}
# Compute seroconversion rate summary and assign to object "seroincidenceSummary".
seroincidenceSummary <- summary(seroincidenceData)

# Show the results.
seroincidenceSummary$Results
```

    ##    Lambda.est Lambda.lwr Lambda.upr  Deviance Convergence Stratum
    ## 1  0.09721890 0.06397004 0.14774909 105.29815           0       1
    ## 2  0.05497303 0.03573489 0.08456817 181.90627           0       2
    ## 3  0.04456096 0.02845903 0.06977327 199.99357           0       3
    ## 4  0.02944038 0.01715302 0.05052964 141.32558           0       4
    ## 5  0.02569781 0.01436524 0.04597053 123.33213           0       5
    ## 6  0.02565297 0.01587275 0.04145941 187.41060           0       6
    ## 7  0.03458616 0.02012722 0.05943207 128.84486           0       7
    ## 8  0.02776873 0.02640621 0.02920154 133.57166          52       8
    ## 9  0.04016261 0.02133209 0.07561541  94.99389           0       9
    ## 10 0.01996477 0.01898409 0.02099612 104.30992          52      10
    ## 11 0.03356017 0.01917771 0.05872886 141.20168           0      11
    ## 12 0.05287648 0.03263923 0.08566140 181.68655           0      12
    ## 13 0.05081097 0.02946963 0.08760729 123.66670          52      13
    ## 14 0.04813104 0.02945312 0.07865371 186.92218           0      14
    ## 15 0.03421042 0.02210895 0.05293569 215.78684           0      15

As one may note it is a dataframe with six columns that can be later post-processed with custom 
calculations.

# 4. Other examples

The following examples show the standard procedure for estimating seroconversion rate.

```{r, eval=FALSE}
# 1. Define cross-sectional data.
serologyData <- with(campylobacterSimMediumData,
                     data.frame(
                       IgG = IgG.ratio.new,
                       IgM = IgM.ratio.new,
                       IgA = IgA.ratio.new))

# 2. Define longitudinal response data
responseParams <- campylobacterDelftParams3

# 3. Define cut-offs
cutOffs <- list(IgG = 0.25, IgM = 0.25, IgA = 0.25)

# 4. Baseline distributions
baseLine <- list(IgG = c(log(0.05), 1),
                 IgM = c(log(0.005), 1),
                 IgA = c(log(0.005), 1))

# 4a. Calculate a single seroincidence rate for all serum samples...
seroincidenceData <- estimateSeroincidence(
  data = serologyData,
  antibodies = c("IgG", "IgM", "IgA"),
  strata = "",
  params = responseParams,
  censorLimits = cutOffs,
  par0 = baseLine)

# 4b. ...or calculate a single seroincidence rate for a single serum sample 
# (triplet of titres)...
seroincidenceData <- estimateSeroincidence(
  data = serologyData[1, ],
  antibodies = c("IgG", "IgM", "IgA"),
  strata = "",
  params = responseParams,
  censorLimits = cutOffs,
  par0 = baseLine)

# 4c. ...or calculate a single seroincidence rate for all serum samples (only IgG)
seroincidenceData <- estimateSeroincidence(
  data = serologyData,
  antibodies = c("IgG"),
  strata = "",
  params = responseParams,
  censorLimits = cutOffs,
  par0 = baseLine)

# 5a. Produce summary of the results with 2.5% and 97.5% bounds...
summary(seroincidenceData)

# 5b. ...or produce summary of the results with 5% and 95% bounds, do not show 
# convergence...
summary(seroincidenceData, quantiles = c(0.05, 0.95), showConvergence = FALSE)

# 5c. ...or produce summary and assign to an object...
seroincidenceSummary <- summary(seroincidenceData)
# ...and work with the results object from now on (here: display the results).
seroincidenceSummary$Results
```

\newpage

# Acknowledgements

Project funded by European Centre for Disease Prevention and Control (ECDC).

# References
