---
title: "Generate a simulated cross-sectional sample and estimate seroincidence"
subtitle: "Enteric Fever using HlyE IgG and/or HlyE IgA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulate_xsectionalData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This vignette shows how to simulate a cross-sectional sample of 
seroresponses for incident infections as a Poisson process with frequency `lambda`. 
Responses are generated for the antibodies given in the `antigen_isos` argument.

Age range of the simulated cross-sectional record is `lifespan`.

The size of the sample is `nrep`.

Each individual is simulated separately, but different antibodies are
modelled jointly (see "sim-graph.r").

Longitudinal parameters are calculated for an age: `age.fx` (fixed age).
However, when `age.fx` is set to NA then the age at infection is used.

The boolean `renew.params` determines whether each infection uses a
new set of longitudinal parameters, sampled at random from the
posterior predictive output of  the longitudinal model. If set to `FALSE`
a parameter set is chosen at birth and kept, but:
1. the baseline antibody levels (`y0`) are updated with the simulated level
(just) prior to infection, and
2. when `is.na(age.fx)` then the selected parameter sample is updated for the
age when infection occurs.

There is also a variable `n.mc`: when `n.mc==0` then a random MC
sample is chosen out of the posterior set (1:4000). When `n.mc` is
given a value in 1:4000 then the chosen number is fixed and reused
in any subsequent infection. This is for diagnostic purposes.


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(serocalculator)
library(tidyverse)


#load in longitudinal parameters, these are modelled from all SEES cases across all ages and countries 
dmcmc <- getAdditionalData(fileURL = "https://osf.io/download/rtw5k")  

```


## Simulate cross-sectional data
```{r}
#set seed to reproduce results
set.seed(54321)

ab.nm <- rbind(c("HlyE","IgA"),c("HlyE","IgG"));
# antibodies <- dmcmc$antigen_iso %>% droplevels() %>% levels()
antibodies = c("HlyE_IgA", "HlyE_IgG")
abs <-  c(1,2);
lambda <- 0.2; # incidence (1/yr)
npar <- 5; # y0, y1, t1, alpha, shape

lifespan <- c(0,20); # range covered in simulations
nrep <- 100; # cross-sectional sample size

dlims <- rbind(c(1e-2,5e-1),c(1e-2,5e-1)); # biologic noise distribution

csdata0 <- sim.cs(
  curve_params = dmcmc,
  lambda = lambda,
  n.smpl = nrep,
  age.rng = lifespan,
  antigen_isos = antibodies,
  n.mc = 0,
  renew.params = TRUE,
  add.noise = TRUE,
  noise_limits = dlims
)



```

## Estimate seroincidence
```{r}

cond <- data.frame(
  antigen_iso = c("HlyE_IgG", "HlyE_IgA"),
  nu = c(0.5, 0.5),  # Biologic noise (nu)
  eps = c(0.2, 0.2),                             # M noise (eps)
  y.low = c(1, 1),                            # low cutoff (llod)
  y.high = c(5e6, 5e6))                      # high cutoff (y.high)


csdataL <- as.data.frame(csdata0) %>%
  pivot_longer(cols= c("HlyE_IgG", "HlyE_IgA"), names_to = "antigen_iso", values_to = "value")

est.incidence(
  csdataL %>%
    rename(y = value,
           a = age), 
  dmcmc, 
  cond, 
  antigen_isos = c("HlyE_IgG", "HlyE_IgA"))


```




# Optional - loop over "clusters" and different lambdas

```{r "parallel"}
library(parallel)
library(doParallel)
n_cores = max(1, parallel::detectCores() - 1)
doParallel::registerDoParallel(cores = n_cores)
nclus = 10;
nrep <- 100; # cross-sectional sample size
lmbdaVec <- c(.05,.1, .15, .2, .3)

sim.df <- 
  foreach(l = lmbdaVec, .combine = bind_rows) %:%
  foreach(n = 1:nclus,  .combine = bind_rows) %dopar% {
    sim.cs(
      lambda = l,
      n.smpl = nrep,
      age.rng = lifespan,
      antigen_isos = antibodies,
      n.mc = 0,
      renew.params = TRUE,
      add.noise = TRUE,
      curve_params = dmcmc,
      noise_limits = dlims,
      baseline_limits = blims
    ) %>% 
      mutate(lambda.sim = l, cluster = n) 
  }
print(sim.df)
```

## Estimate incidence in each cluster

```{r, "est-by-stratum"}
ests = 
  sim.df %>% 
  as_tibble() %>%
  pivot_longer(
    cols = c("HlyE_IgG", "HlyE_IgA"), 
    names_to = "antigen_iso", 
    values_to = "value") %>%
  rename(
    y = value,
    a = age) %>% 
  est.incidence.by(
    curve_params = dmcmc,
    noise_params = cond,
    numCores = n_cores,
    strata = c("lambda.sim", "cluster"),
    curve_strata_varnames = NULL,
    noise_strata_varnames = NULL,
    antigen_isos = c("HlyE_IgG", "HlyE_IgA")
  )

summary(ests)
```

## plot distribution of estimates by simulated incidence rate

```{r, "graph", fig.width = 7}

library(ggplot2)
summary(ests) |> 
  ggplot(aes(x = lambda.sim, y = incidence.rate)) +
  geom_point() +
  xlab("Simulated incidence rate") +
  ylab("Estimated incidence rate") +
  theme_bw() +
  geom_abline(aes(intercept = 0, slope = 1, col = "x=y")) +
  expand_limits(x = 0,y = 0) +
  labs(col = "")
```

