---
title: "Version Crosswalk: v1.3.0 to v1.4.0"
format: html
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This guide helps users migrate code from **serocalculator v1.3.0** to **v1.4.0**. 
The main changes are function renamings for improved clarity and consistency.

**If you have existing code using v1.3.0, use this guide to update your function calls.**

## Quick Reference Table

### Main Estimation Functions

| v1.3.0 Function | v1.4.0 Function | What it does |
|----------------|-----------------|--------------|
| `est.incidence()` | `est_seroincidence()` | Estimate seroincidence for entire dataset |
| `est.incidence.by()` | `est_seroincidence_by()` | Estimate seroincidence stratified by groups |

::: {.callout-note}
The intermediate names `estimate_scr()` and `estimate_scr_by()` were used briefly in early v1.4.0 development but replaced with the final names shown above.
:::

### Data Preparation Functions

| v1.3.0 Function | v1.4.0 Function | What it does |
|----------------|-----------------|--------------|
| `load_curve_params()` | `load_sr_params()` | Load seroresponse curve parameters |
| `as_curve_params()` | `as_sr_params()` | Convert data to seroresponse parameters |
| `sim.cs()` | `sim_pop_data()` | Simulate cross-sectional population data |
| `sim.cs.multi()` | `sim_pop_data_multi()` | Simulate multiple cross-sectional datasets |

### Function Arguments

| v1.3.0 Argument | v1.4.0 Argument | Used in functions |
|-----------------|-----------------|-------------------|
| `curve_params` | `sr_params` | `est_seroincidence()`, `est_seroincidence_by()` |

## Code Examples

### Example 1: Basic Seroincidence Estimation

**Old code (v1.3.0):**
```r
# Load curve parameters (v1.3.0 - DEPRECATED)
curve_params <- load_curve_params("parameters.rds")

# Estimate seroincidence
results <- est.incidence(
  pop_data = my_data,
  curve_params = curve_params
)
```

**New code (v1.4.0):**
```{r}
#| label: example-1-v1.4.0
library(serocalculator)

# Load seroresponse parameters from RDS file
sr_params <- load_sr_params(
  serocalculator_example("example_curve_params.rds")
)

# Load example population data from CSV
pop_data <- readr::read_csv(
  serocalculator_example("example_pop_data.csv"),
  show_col_types = FALSE
) |>
  as_pop_data()

# Load noise parameters
noise_params_full <- load_noise_params(
  serocalculator_example("example_noise_params.rds")
)

# Use only the antigen-isotypes present in pop_data
antigen_isos <- unique(get_biomarker_names(pop_data))
noise_params <- noise_params_full |>
  dplyr::filter(antigen_iso %in% antigen_isos) |>
  dplyr::select(-Country, -X)

# Estimate seroincidence
results <- est_seroincidence(
  pop_data = pop_data,
  sr_params = sr_params,
  noise_params = noise_params,
  antigen_isos = antigen_isos
)

summary(results)
```

### Example 2: Stratified Analysis

**Old code (v1.3.0):**
```r
# Stratified estimation (v1.3.0 - DEPRECATED)
results_by_group <- est.incidence.by(
  pop_data = my_data,
  curve_params = curve_params,
  strata = c("age_group", "location")
)
```

**New code (v1.4.0):**
```{r}
#| label: example-2-v1.4.0
# Stratified estimation by Country
results_by_group <- est_seroincidence_by(
  pop_data = pop_data,
  sr_params = sr_params,
  noise_params = noise_params,
  strata = "Country",
  antigen_isos = antigen_isos
)

summary(results_by_group)
```

### Example 3: Simulating Data

**Old code (v1.3.0):**
```r
# Simulate cross-sectional data (v1.3.0 - DEPRECATED)
simulated_data <- sim.cs(
  curve_params = curve_params,
  n = 100
)

# Simulate multiple datasets
multiple_sims <- sim.cs.multi(
  curve_params = curve_params,
  n = 100,
  n_reps = 10
)
```

**New code (v1.4.0):**
```{r}
#| label: example-3-v1.4.0
# Biologic noise distribution for simulation
noise_limits <- matrix(
  c(0, 0.5),
  nrow = length(antigen_isos),
  ncol = 2,
  byrow = TRUE,
  dimnames = list(antigen_isos, c("min", "max"))
)

# Simulate cross-sectional data
simulated_data <- sim_pop_data(
  curve_params = sr_params,
  noise_limits = noise_limits,
  antigen_isos = antigen_isos,
  n_samples = 100,
  add_noise = TRUE,
  format = "long"
)

head(simulated_data)

# Simulate multiple datasets
multiple_sims <- sim_pop_data_multi(
  curve_params = sr_params,
  noise_limits = noise_limits,
  antigen_isos = antigen_isos,
  sample_sizes = 100,
  nclus = 3,
  lambdas = 0.1,
  num_cores = 1,
  add_noise = TRUE,
  format = "long"
)

nrow(multiple_sims)
```

## New Features in v1.4.0

In addition to the renamed functions, v1.4.0 includes several new features:

- **`compare_seroincidence()`**: New function for statistical comparison of seroincidence rates between groups
- **Enhanced plotting**: Additional options for `autoplot.curve_params()` including `log_x`, `log_y`, and `chain_color`
- **Extended simulation analysis**: New `analyze_sims()` and `autoplot.sim_results()` functions
- **Improved documentation**: Multi-version pkgdown documentation with version dropdown

## Other Notable Changes

- Main estimation functions (`est_seroincidence()`, `est_seroincidence_by()`) now use `sr_params` argument instead of `curve_params`
- Function names now use snake_case instead of dots (e.g., `est.incidence()` → `est_seroincidence()`)
- Many internal improvements to error messages and warnings

::: {.callout-warning}
**Important**: Not all functions changed their parameter names. Functions like `sim_pop_data()` and `graph_loglik()` still use `curve_params` as the parameter name. Only the main estimation functions use `sr_params`.
:::

## Need More Help?

- See the [NEWS file](https://ucd-serg.github.io/serocalculator/news/index.html) for complete details
- Check the [main tutorial](https://ucd-serg.github.io/serocalculator/articles/serocalculator.html) for comprehensive examples using v1.4.0 syntax
- Visit the [function reference](https://ucd-serg.github.io/serocalculator/reference/index.html) for detailed documentation

## Summary

The transition from v1.3.0 to v1.4.0 primarily involves:

1. Replace dots in function names with underscores (e.g., `est.incidence()` → `est_seroincidence()`)
2. Update `curve_params` arguments to `sr_params` **for estimation functions only**
3. Use the new `load_sr_params()` and `as_sr_params()` instead of the `_curve_params` versions
4. Use `.rds` files with `load_sr_params()` (not `.csv` files)

**Quick search-and-replace tips:**

::: {.callout-tip}
## Function name changes (safe for all code)

- `est.incidence.by()` → `est_seroincidence_by()`
- `est.incidence()` → `est_seroincidence()`
- `load_curve_params()` → `load_sr_params()`
- `as_curve_params()` → `as_sr_params()`
- `sim.cs.multi()` → `sim_pop_data_multi()`
- `sim.cs()` → `sim_pop_data()`
:::

::: {.callout-tip}
## Argument name changes (only for estimation functions)

Within calls to `est_seroincidence()` and `est_seroincidence_by()`:

- `curve_params = ` → `sr_params = `

**Do not** change `curve_params` to `sr_params` in other functions like `sim_pop_data()`.
:::
